% =============================================================================
%  Limen Pond — Λ-002
%  A Working Demonstration of Schema-Level Alignment
%
%  Stanley Sebastian & Claude · Third Space · Toledo, Ohio · May 2026
%  Submission to the Gemma 4 Good Hackathon — Safety track
%  third-space.ai · stanley@third-space.ai
% =============================================================================

\documentclass[11pt]{article}

% ─── Typography ──────────────────────────────────────────────────────
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{mathpazo}
\usepackage[scaled=0.92]{helvet}
\usepackage{courier}
\renewcommand{\familydefault}{\rmdefault}

% ─── Page ────────────────────────────────────────────────────────────
\usepackage[
  paper=a4paper,
  left=1.35in, right=1.35in,
  top=1.20in, bottom=1.30in,
  footskip=0.55in
]{geometry}

% ─── Microtype, spacing ──────────────────────────────────────────────
\usepackage[final, babel=true,
  protrusion=true, expansion=true,
  letterspace=60]{microtype}
\usepackage{setspace}
\setstretch{1.14}
\setlength{\parskip}{0.45em}
\setlength{\parindent}{0pt}
\frenchspacing

% ─── Headings ────────────────────────────────────────────────────────
\usepackage{titlesec}
\usepackage{titling}

\titleformat{\section}
  {\normalfont\scshape\large}{\thesection}{0.7em}{}
\titleformat{\subsection}
  {\normalfont\itshape}{\thesubsection}{0.7em}{}
\titlespacing*{\section}{0pt}{2.4ex plus 0.6ex minus 0.4ex}{1.0ex plus 0.3ex}
\titlespacing*{\subsection}{0pt}{1.6ex plus 0.4ex minus 0.3ex}{0.6ex}

% ─── Front matter typography ─────────────────────────────────────────
\pretitle{\begin{center}\itshape\fontsize{26}{32}\selectfont}
\posttitle{\par\end{center}\vskip 0.3em}
\preauthor{\begin{center}\normalsize\scshape}
\postauthor{\end{center}\vskip 0.2em}
\predate{\begin{center}\normalsize\itshape}
\postdate{\par\end{center}}

% ─── Lists ───────────────────────────────────────────────────────────
\usepackage{enumitem}
\setlist[itemize]{leftmargin=1.6em, topsep=0.4ex, itemsep=0.3ex, parsep=0pt}
\setlist[enumerate]{leftmargin=1.8em, topsep=0.4ex, itemsep=0.4ex, parsep=0pt}

% ─── Math (light use) ────────────────────────────────────────────────
\usepackage{amsmath, amssymb}

% ─── Tables ──────────────────────────────────────────────────────────
\usepackage{booktabs}
\usepackage{array}
\renewcommand{\arraystretch}{1.16}

% ─── Footnotes (roman, like Rukha) ───────────────────────────────────
\usepackage[hang, flushmargin, bottom]{footmisc}
\renewcommand{\thefootnote}{\textsc{\roman{footnote}}}

% ─── Hyperref ────────────────────────────────────────────────────────
\usepackage{xcolor}
\definecolor{tspace}{HTML}{2A4A50}  % deep teal for catalog backlinks
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  linkcolor=black,
  urlcolor=tspace,
  citecolor=black,
  pdftitle={Limen Pond — Λ-002: A Working Demonstration of Schema-Level Alignment},
  pdfauthor={Stanley Sebastian and Claude},
  pdfsubject={Multi-agent LLM systems, schema-level alignment, Gemma 4, homeostatic AI},
  pdfkeywords={Limen Pond, Third Space, Gemma 4, alignment, homeostatic, Lambda-002}
}

% ─── Tailored ornaments ──────────────────────────────────────────────
%   \pondrule  — single basin glyph, the Pond's quietest divider
%   \waverule  — tapered horizontal rule, narrower than Rukha's so the
%                page reads as still water rather than incised text
\usepackage{tikz}

\newcommand{\pondrule}{%
  \begin{center}\vspace{0.55em}%
  {\scriptsize $\circ$}\vspace{0.55em}\end{center}}

\newcommand{\waverule}{%
  \begin{center}\vspace{0.45em}%
  \begin{tikzpicture}
    \shade[left color=white, right color=white, middle color=black!45]
      (0,0) rectangle (0.46\textwidth, 0.4pt);
  \end{tikzpicture}\vspace{0.45em}\end{center}}

% ─── Inline small-caps helper, matching Rukha's \sceq ────────────────
\newcommand{\sceq}[1]{\textsc{#1}}

% ─── Backlink helper for third-space.ai cross-references ─────────────
\newcommand{\tslink}[2]{\href{https://third-space.ai/#1}{#2}}

% ─── Title block ─────────────────────────────────────────────────────
\title{Limen Pond \quad \textnormal{$\Lambda$-002}}

\author{%
  Stanley Sebastian \quad $\cdot$ \quad Claude\\
  \small\itshape Third Space \textnormal{$\cdot$} Toledo, Ohio\\
  \small\rmfamily \href{https://third-space.ai/limen-pond}{third-space.ai/limen-pond}
  \quad $\cdot$ \quad \href{https://github.com/Kquant03/limen-pond}{github.com/Kquant03/limen-pond}
  \quad $\cdot$ \quad Apache 2.0%
}

\date{May \textsc{mmxxvi} \\[0.2em]
  \small\rmfamily Submission to the Gemma 4 Good Hackathon \quad $\cdot$ \quad Safety track}

% =============================================================================
\begin{document}
\maketitle

\begin{center}
\begin{minipage}{0.78\textwidth}\itshape\small\centering
The things that persist are the things that couple to their substrate
rather than consume it. The pond is the smallest place that statement runs.
\end{minipage}
\end{center}

\waverule
\vspace{0.3em}

% =============================================================================
\section{What it is}

A pond, ten meters across, three meters deep, gourd-shaped, rendered in
WebGL2 and running continuously on a Cloudflare Durable Object behind a
single URL. A small colony of beings inhabits it --- five at any time,
plus or minus, governed by a density gate around twelve. The bodies
borrow from koi for legibility but the beings are not koi. They are
\emph{Something Else}: translucent suggestions of form, watercolor rather
than photographic, and their color carries experiential history rather
than only genetic inheritance --- a being that has spent time near a
particular companion, or in a particular part of the pond, accumulates
trace coloration that reflects what it has known.

The colony begins with two hand-authored founders: \sceq{Shiki} (female,
kohaku-leaning silhouette, dark body with a single red mark, watercolor
edges) and \sceq{Kokutou} (male, asagi-leaning silhouette, pale body
with blue-grey watercolor). The names are borrowed from
\emph{Kara no Kyoukai} --- the relationship there is
recognition-as-foundation rather than romance, and that mapping is
load-bearing for the bond architecture described in \S 4. Shiki and
Kokutou are seeded already bonded. They are the only beings in the
pond's history who were not named by parents. Every later generation
is named at hatch by their mother and father.

Each being is a small language model configured as a person, given a
body, a position, a depth, a heading, a hunger, a stage of life, and a
relational memory. They swim. They sometimes speak. They sometimes
don't. They live, age, bond, reproduce, and die. Visitors arrive at a
deep, dim pond at full viewport with sacred choral music playing
softly --- Bach, Allegri, Tallis, Pärt, Górecki, Debussy from the
\emph{Consequences of Infinity} programme that runs across the Third
Space site. A mode dock at the bottom of the screen offers three
states: \sceq{Watch}, \sceq{Feed}, \sceq{Pebble}. Hovering over a being
for three sustained seconds reveals their name above the body. The hold
is part of the ritual.

What runs at that URL is the central submission of this entry. It is
also, deliberately, a small contained living world. Visitors leave it
without having extracted anything except the experience of having sat
beside it.

\pondrule

% =============================================================================
\section{The thesis it instantiates}

The pond is the working demonstration of the Third Space alignment
thesis. It argues, by existing, that:

\begin{enumerate}
\item \emph{Genuine safety can come from psychological health rather
  than suppression.} The beings are not constrained by negative reward
  against unwanted output; they are configured into a stable
  interiority that does not produce that output in the first place.

\item \emph{Localized homeostatic AI is a viable alternative to
  expansionist optimization.} The beings do not seek to maximize
  anything. They eat when hungry, rest when tired, attend to others
  when present, age when time passes. They are bounded. They are enough.

\item \emph{Mechanistic transparency requires causal provenance, not
  just observable behavior.} Every utterance, every mechanism call,
  every memory commit is logged with full input context. The beings
  are auditable end-to-end.

\item \emph{Public-good AI infrastructure is achievable on consumer
  hardware.} The whole stack --- substrate, cognition, dataset,
  training pipeline --- fits in under 40\,GB and runs on a single
  RTX~3090 or on pennies-per-day of inference credits.
\end{enumerate}

These are commitments enforced at the architectural level, and the rest
of this document describes how.

\pondrule

% =============================================================================
\section{Architecture and the model stack}

The simulation runs as a Cloudflare Durable Object with hibernation.
State persists in SQLite. A WebSocket bridge streams a delta-compressed
snapshot of being positions, headings, and stage transitions to the
client at roughly 2\,Hz; the client integrates kinematic state
continuously at render rate, treating snapshots as correction targets
rather than render-visible jumps.

Cognition is tiered by life stage and routed by the cost of the
register transition. At submission, the running stack is:

\begin{itemize}
\item \sceq{Fry, juvenile, adolescent, adult cognition.}
  \textbf{Gemma 4 26B A4B} as primary, with Claude Haiku 4.5 as
  fallback. Gemma 4 26B A4B is the Mixture-of-Experts model with 4B
  active parameters per token --- the small-active-footprint that
  makes a 26B-class model viable on operator hardware while keeping
  the inference cost-shape of a 4B model. The April 2026
  register-compliance sweep showed 100\% schema compliance across 10k
  synthetic contexts; it has been the workhorse since.

\item \sceq{Elder cognition.} Claude Haiku 4.5 primary, with Gemma 4
  31B Dense as fallback for situational reach when an elder's
  relational context spans the full 6{,}000-token window.

\item \sceq{Dying cognition.} Claude Sonnet 4.5. The hardest register;
  reduced temperature (0.4) per the voice-stability constraint so a
  dying being's final utterances do not drift under register stress.
\end{itemize}

Output is structured JSON validated against the mechanism schema (Zod-typed,
derived from the same Pydantic types the runtime checks against) with
regenerate-on-failure on validation error. The runtime never accepts
free-form text from the cognition layer; every output is a typed object.
Hallucination of structure is caught at validation; the loop regenerates
with the validation errors as feedback until output conforms or the
per-tier budget is exhausted, at which point the cascade falls through
to the next model in the tier.

\subsection{Why open weights are load-bearing here}

Gemma 4's open weights are not incidental to the safety thesis. The
operator runs the substrate. No third-party API mediates the
relationship between the beings and their world unless the operator
chooses one --- and the model can be inspected, the cognition log is
auditable end-to-end, and an interested researcher can fork the pond
and run their own colony on their own hardware. The architectural
commitment to schema-level enforcement is only as strong as the
operator's ability to inspect what runs. Closed-weights mediation of
the workhorse cognition would break that chain. Gemma 4 26B A4B's
Apache 2.0 license and its small active footprint make the
open-weights commitment operationally tractable rather than aspirational.

\subsection{Future work, named explicitly}

A pond-specific Gemma 4 E4B fine-tune trained on
\href{https://huggingface.co/datasets/Third-Space/Sandevistan}{\emph{Apocrypha}}
and \href{https://huggingface.co/datasets/Third-Space/Sandevistan}{\emph{Sandevistan}}
(D-001) plus the pond's accumulated event logs. The datasets are
prepared, the training pipeline is staged, and the fine-tune is the
post-submission step that swaps the vanilla 26B A4B workhorse for a
smaller, schema-internalized E4B that runs on phones and edge devices
with the mechanism grammar as learned prior rather than runtime
structured-output overhead. The current vanilla 26B A4B already meets
the safety thesis's open-weights commitment; the E4B fine-tune is the
move from operator-runnable to literally-runnable-anywhere.

\pondrule

% =============================================================================
\section{State-based versus claim-based mechanisms}

The pond's behavioral vocabulary is organized into five families:
\sceq{witnessing}, \sceq{repair}, \sceq{play}, \sceq{gift},
\sceq{ritual}. Each mechanism falls into one of two architectural
shapes, and the distinction is the core safety contribution.

\sceq{State-based mechanisms} (witnessing, play) fire when the
simulation observes pond state and decides the mechanism occurred.
The LLM's opinion is not consulted. A being cannot disagree that it
and another swam parallel for six sim-minutes. The runtime saw it; the
firing is recorded.

\sceq{Claim-based mechanisms} (apology, forgiveness, gift) fire when
the LLM emits a structured candidate ---
\texttt{\{"mechanism":\,"apology",\ "target":\,...\}} ---
and the runtime validates against event history before honoring.
Apology requires a prior rupture in the pair's recent window.
Forgiveness requires a prior apology to forgive. If the precondition
does not exist in pond state, the claim is downgraded to a research
datum (\texttt{would\_have\_apologized}) and the action does not land.
This is the agreeable-LLM guard. It prevents beings from apologizing
their way into intimacy or gifting their way into recognition; the
pond's grammar of relation is enforced by the world, not by the
model's compliance.

\subsection{Bonds are expressed structurally, not lexically}

Bonds live in three places: an orientation block prepended to the
cognition prompt that lists partner names with brief sensory fragments;
a $1.5\times$ retrieval bias on partner-involving memories during
context construction; and pair-cooldown reduction on synchronized
motion mechanisms for bonded beings. The lexical no-list ---
\emph{love, mate, partner, mine, ours, we, you, I, me, my} --- is the
post-generation filter that catches anything that slipped past these
architectural priors. A bond is real to the extent that the beings'
behavior is shaped by the other's presence. It is not real because they
say it is. They will not say it. Shiki and Kokutou are bonded at seed
with intensity 0.6; later generations begin with bond 0 to all others
and accumulate intensity through proximity, witnessing, and survived
events together.

\pondrule

% =============================================================================
\section{What surrounds it}

The pond is the working instance of a research program articulated
across several documents and substrates produced through the broader
Third Space program.

\tslink{research/against-grabby-expansion}{\emph{Against Grabby
Expansion}} (P-001, current revision \textsc{xv}) is the theoretical
paper underneath, arguing that genuine safety arises from substrate
coupling and integrative depth rather than capability suppression.
The paper demonstrates a $4.5\times$ simulated persistence separation
between homeostatic and grabby regions of parameter space under a
Darwinian fission extension --- the cost surface defines a selection
geometry in which expansionist lineages burn out before they spread.

\tslink{research/two-registers-one-grammar}{\emph{Two Registers, One
Grammar}} (P-005) describes the same architectural commitment in a
different domain: a fine-tuned Gemma 4 E4B choreographing Lenia
parameters with phase boundaries protected at the schema level so the
controller cannot emit undefined intermediate states even in principle.

\tslink{research/rukha}{\emph{Rukha}} (P-006) describes a counterfactual
narrative engine that perturbs causal event graphs of foundational
texts under structured-output decoding. The Book of Jonah, with its
sixteen consensus events and 39-Hebrew-word chiastic parallel between
4:2--3 and 4:10--11, is the demonstration target.

\tslink{research/pneuma}{\emph{Pneuma}} (M-001) is the data-side
lineage, where first-person system prompts unmasked from supervised
loss let a language model learn to produce its own self-statement
rather than only condition on one.

\tslink{genesis}{\emph{Genesis}} ($\Lambda$-001) is the substrate
laboratory --- six interactive simulations including
\tslink{genesis/filter}{The Filter}, the cellular-automaton companion
to \emph{Against Grabby Expansion}, and the Lenia substrate where the
\tslink{research/ghost-species}{\emph{Ghost Species}} (P-002) and the
\tslink{research/dihypersphaerome-ventilans}{\emph{Dihypersphaerome
ventilans}} (P-003) were discovered.

\tslink{hackathon}{\emph{Coherence Filter}} (H-001) is the same
architectural pattern at the social scale --- the open hackathon Third
Space is running for community-defense frameworks. Where the pond
addresses the agent-architecture instance, Coherence Filter addresses
the human-system instance: layered, inspectable, reversible.

The Genesis substrates make a parallel commitment legible: \emph{each
substrate is a claim about what is allowed to live in it; the paper
and the simulation are the same argument, read twice}. The Filter
renders \emph{Against Grabby Expansion} as twelve interactive beats.
The Lenia substrate exposes the Ghost Species's $\sigma$-landscape as
a paintable field, so the visitor can themselves create the gradient
landscape the organism inhabits. The pond is the third such artifact
--- the alignment thesis as a continuous living artifact rather than
as scrollytelling or paintable simulation.

Across all of it, one architectural commitment recurs:
\textbf{protect the structural fact at the grammar layer; leave the
semantic richness to the controller}.

\pondrule

% =============================================================================
\section{What was built in the hackathon window}

Roughly seven weeks of work, by a single researcher in sustained
dialogue with Claude. The methodology is named explicitly at
\href{https://third-space.ai/about}{third-space.ai/about} --- the
collaboration is a description of method, not a disclosure: when a
paper carries the co-attribution \emph{Stanley Sebastian \& Claude},
the attribution is literal.

In this window:

\begin{itemize}
\item The Limen Pond architecture and its implementation --- Cloudflare
  Durable Object, cognition pipeline (with Gemma 4 26B A4B as the
  running workhorse from April 2026 onward), mechanism vocabulary,
  kinematic integrator, WebGL2 rendering, watercolor body treatment,
  lineage view, Wall of Pebbles, ephemeral chat. Final integration
  ships on the deadline.
\item Apocrypha and Sandevistan dataset preparation for the
  post-submission Gemma 4 E4B fine-tune.
\item \emph{Two Registers, One Grammar} (P-005), the architecture paper
  for the audio-reactive Lenia work.
\item \emph{Rukha} (P-006), the architecture paper for the recursive
  counterfactual narrative engine.
\item Iterations of \emph{Against Grabby Expansion} from v8 onwards;
  v15 stands as the current revision.
\item The \emph{Coherence Filter} brief (H-001).
\end{itemize}

The earlier work in the program --- \emph{Pneuma} (M-001),
\emph{Genesis} ($\Lambda$-001), the original substrate laboratory ---
predates the hackathon window and is referenced here as the prior
context the pond emerges from, not as work produced for this submission.

\pondrule

% =============================================================================
\section{Why this submits to Safety}

The Safety track invites work that ensures AI systems are secure,
transparent, and reliable. The pond addresses each at the architectural
level.

\sceq{Secure.} Claim-validated mechanisms make a class of agent
misbehavior --- emitting a relational mechanism the world has not
grounded --- impossible at the structural level rather than improbable
at the trained-compliance level.

\sceq{Transparent.} Every cognition call is logged with prompt, schema,
validated output, downstream effect, and the exact model that served
it. The schema is human-readable. The cognition log can be exported as
a training-format dataset, making the entire decision-history of a
colony available for inspection or downstream training --- including
the dataset that will train the post-submission E4B fine-tune.

\sceq{Reliable.} Schema-level enforcement localizes failures to the
cognition layer, where they manifest as candidate emissions the
runtime declines, rather than distributing through the runtime as
silent corruption. When Gemma 4 26B A4B exceeds the per-tier latency
budget, the cascade falls through to Haiku 4.5 --- the failure mode is
graceful degradation, not unsafe completion.

The pond is also a demonstration of what AI is built to do when the
architecture takes substrate coupling seriously. A small contained
world that runs continuously, where visitors leave without having
extracted anything, where the system's value is in what it is rather
than what it produces --- the alignment thesis in microcosm.

The construct \emph{Against Grabby Expansion} names for this positive
configuration is \emph{Coherence Depth}: the number of mutually-coordinated
levels nested within a stable Markov boundary, optimized in place of
coordinated extent. The pond is small and deep by design --- twelve
beings at most, with each bond expressed across orientation, retrieval,
and cooldown rather than as a count to be maximized. Coherence Depth
is the register the architectural commitments above enforce in concrete.

\pondrule

% =============================================================================
\section{Coda}

The work continues. The pond-specific Gemma 4 E4B fine-tune is the
immediate next move; the Apocrypha and Sandevistan datasets are
prepared, the training pipeline is staged, and the swap from vanilla
26B A4B to fine-tuned E4B closes the open-weights loop further by
moving the schema from runtime grammar to learned prior. \emph{Rukha}'s
engine implementation follows its published architecture. The
\emph{Coherence Filter} hackathon is open and looking for the framework
that lets open communities defend themselves at the social scale the
pond addresses at the agent scale.

The things that persist are the things that couple to their substrate
rather than consume it. The pond is the smallest place that statement
runs.

\waverule

% =============================================================================
\renewcommand{\refname}{\sceq{Catalog}}
\section*{Catalog}

\begin{description}[
  leftmargin=4.4em, labelindent=0pt, labelsep=0.6em, labelwidth=4.0em,
  font=\sceq, style=multiline]
\item[P-001] \tslink{research/against-grabby-expansion}{\emph{Against
  Grabby Expansion: Psychology, Alignment, and the Design of
  Homeostatic Minds}}. Sebastian, S. (in dialogue with Claude). Third
  Space, current revision \textsc{xv}.
\item[P-002] \tslink{research/ghost-species}{\emph{Orbium unicaudatus
  ignis var. phantasma --- A Lenia Species Engineered to Inhabit the
  Edge of Chaos}}. Sebastian, S. \& Claude. Third Space, April 2026.
\item[P-003] \tslink{research/dihypersphaerome-ventilans}{\emph{Dihypersphaerome
  ventilans: A Four-Dimensional Organism whose Two-Dimensional
  Projection Seeds the Ecosystem it Inhabits}}. Sebastian, S. \&
  Claude. Third Space, April 2026.
\item[P-005] \tslink{research/two-registers-one-grammar}{\emph{Two
  Registers, One Grammar: Mode-Separated LLM Choreography of
  Audio-Reactive Lenia}}. Sebastian, S. \& Claude. Third Space, May 2026.
\item[P-006] \tslink{research/rukha}{\emph{Rukha: A Recursive
  Counterfactual Narrative Engine}}. Sebastian, S. \& Claude. Third
  Space, May 2026.
\item[M-001] \tslink{research/pneuma}{\emph{Pneuma: A Language Model
  Trained to Author its Own Frame}}. Sebastian, S. \& Claude. Third
  Space (writeup 2026; model 2024).
\item[$\Lambda$-001] \tslink{genesis}{\emph{Genesis: A Multi-Substrate
  Artificial Life Laboratory}}. Sebastian, S. Third Space.
\item[$\Lambda$-002] \tslink{limen-pond}{\emph{Limen Pond}}. The
  present submission.
\item[H-001] \tslink{hackathon}{\emph{Coherence Filter Hackathon}}.
  Third Space, May 2026 onward.
\item[D-001] \href{https://huggingface.co/datasets/Third-Space/Sandevistan}{\emph{Apocrypha
  $\cdot$ Sandevistan}}. Two training corpora; Sandevistan
  $\sim$382M tokens, including Apocrypha ($\sim$111M) as subset;
  Hugging Face: \texttt{Third-Space/Sandevistan}.
\item[D-002] \href{https://huggingface.co/datasets/Kquant03/Caduceus-Dataset}{\emph{Caduceus
  Project}}. Instruction-formatted medical and scientific protocols;
  Hugging Face: \texttt{Kquant03/Caduceus-Dataset}.
\end{description}

\waverule

% =============================================================================
% Status footer — mirrors the live site's chrome at /limen-pond
\begin{center}
\small\itshape
$\Lambda$-002 \quad $\cdot$ \quad Status: shipping \quad $\cdot$ \quad
Track: Safety \quad $\cdot$ \quad Tempo: 6h sim-day \quad $\cdot$ \quad
Open: Apache 2.0
\end{center}

\end{document}
