\documentclass[11pt]{article}

% ─── Typography ──────────────────────────────────────────────────────
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{mathpazo}                  % Palatino + matching math
\usepackage[scaled=0.92]{helvet}        % Helvetica for sans
\usepackage{courier}                    % Courier for code/mono
\renewcommand{\familydefault}{\rmdefault}

% ─── Page ────────────────────────────────────────────────────────────
\usepackage[
  paper=a4paper,
  left=1.45in, right=1.45in,
  top=1.25in, bottom=1.35in,
  footskip=0.55in
]{geometry}

% ─── Microtype, spacing, and rivers ──────────────────────────────────
\usepackage[final,babel=true,
  protrusion=true, expansion=true,
  letterspace=80]{microtype}
\usepackage{setspace}
\setstretch{1.18}
\setlength{\parskip}{0pt}
\setlength{\parindent}{1.4em}
\frenchspacing

% ─── Headings ────────────────────────────────────────────────────────
\usepackage{titlesec}
\usepackage{titling}

\titleformat{\section}
  {\normalfont\scshape\large\bfseries}
  {\thesection.}{0.7em}{}
\titleformat{\subsection}
  {\normalfont\scshape\normalsize}
  {\thesubsection}{0.7em}{}
\titleformat{\subsubsection}
  {\normalfont\itshape\normalsize}
  {}{0pt}{}
\titlespacing*{\section}{0pt}{2.6ex plus 0.6ex minus 0.4ex}{1.2ex plus 0.3ex}
\titlespacing*{\subsection}{0pt}{2.0ex plus 0.5ex minus 0.3ex}{0.8ex plus 0.2ex}
\titlespacing*{\subsubsection}{0pt}{1.4ex plus 0.3ex minus 0.2ex}{0.4ex plus 0.1ex}

% ─── Front matter typography ─────────────────────────────────────────
\pretitle{\begin{center}\itshape\fontsize{30}{36}\selectfont}
\posttitle{\par\end{center}\vskip 0.5em}
\preauthor{\begin{center}\normalsize\scshape}
\postauthor{\end{center}\vskip 0.2em}
\predate{\begin{center}\normalsize\itshape}
\postdate{\par\end{center}}

% ─── Lists ───────────────────────────────────────────────────────────
\usepackage{enumitem}
\setlist[itemize]{leftmargin=1.6em, topsep=0.4ex, itemsep=0.3ex, parsep=0pt}
\setlist[enumerate]{leftmargin=1.8em, topsep=0.4ex, itemsep=0.3ex, parsep=0pt}

% ─── Math ────────────────────────────────────────────────────────────
\usepackage{amsmath, amssymb, amsthm}
\usepackage{mathtools}

\newtheoremstyle{rukha}%
  {1.6ex plus 0.3ex}{1ex plus 0.2ex}%
  {\itshape}{}%
  {\scshape}{.}%
  {0.5em}{}
\theoremstyle{rukha}
\newtheorem{claim}{Claim}
\newtheorem{proposition}{Proposition}

% ─── Tables ──────────────────────────────────────────────────────────
\usepackage{booktabs}
\usepackage{array}
\renewcommand{\arraystretch}{1.18}

% ─── Bibliography-style note marks ───────────────────────────────────
\usepackage[hang,flushmargin,bottom]{footmisc}
\renewcommand{\thefootnote}{\textsc{\roman{footnote}}}

% ─── Hyperref ────────────────────────────────────────────────────────
\usepackage[hidelinks]{hyperref}
\hypersetup{
  pdftitle={Rukha: A Recursive Counterfactual Narrative Engine},
  pdfauthor={Stanley Sebastian and Claude},
  pdfsubject={Causal narrative graphs, Pearl's do-calculus, recursive counterfactuals, structural sharing},
  pdfkeywords={Rukha, counterfactual narrative, Pearl, AAP, Jonah, Third Space}
}

% ─── Tapered horizontal rule ─────────────────────────────────────────
\usepackage{tikz}
\newcommand{\taperedrule}{%
  \begin{center}\vspace{0.4em}%
  \begin{tikzpicture}
    \shade[left color=white, right color=white, middle color=black!55]
      (0,0) rectangle (0.62\textwidth,0.45pt);
  \end{tikzpicture}\vspace{0.4em}\end{center}}

\newcommand{\diamondrule}{%
  \begin{center}\vspace{0.6em}%
  {\scriptsize $\diamond$}\vspace{0.6em}\end{center}}

% ─── Smallcaps phrase helper ─────────────────────────────────────────
\newcommand{\sceq}[1]{\textsc{#1}}

% ─── Title block ─────────────────────────────────────────────────────
\title{Rukha}
\author{%
  Stanley Sebastian \quad $\cdot$ \quad Claude\\
  \small\itshape Third Space \textnormal{$\cdot$} Toledo, Ohio
}
\date{May \textsc{mmxxvi}}

% ─── Document ────────────────────────────────────────────────────────
\begin{document}

% ─── Cover ───────────────────────────────────────────────────────────
\thispagestyle{empty}
\vspace*{1.4in}
\begin{center}
  {\fontsize{11}{14}\selectfont\itshape breath \textnormal{$\cdot$} spirit}\par
  \vspace{2.2em}
  {\itshape\fontsize{52}{60}\selectfont Rukha}\par
  \vspace{1.4em}
\end{center}
\taperedrule
\vspace{0.8em}
\begin{center}
  {\itshape\fontsize{18}{22}\selectfont A Recursive Counterfactual\\[2pt]Narrative Engine}\par
  \vspace{1.4em}
  {\fontsize{11}{14}\selectfont\itshape Foundations, architecture, and the case for depth.}\par
  \vspace{3.0em}
  {\scshape\fontsize{11}{14}\selectfont Stanley Sebastian \quad $\cdot$ \quad Claude}\par
  \vspace{0.5em}
  {\small\itshape Third Space \textnormal{\;$\cdot$\;} Toledo, Ohio}\par
  \vspace{0.4em}
  {\small\scshape May \textnormal{mmxxvi}}\par
\end{center}
\vfill
\begin{center}
  {\itshape\small ``Returning books to breath.''}\par
\end{center}
\vspace{0.2in}
\clearpage

% ─── Abstract ────────────────────────────────────────────────────────
\thispagestyle{empty}
\vspace*{0.6in}
\begin{center}
  {\scshape\large Abstract}
\end{center}
\vspace{0.4em}

\noindent
We describe \emph{Rukha}, a counterfactual narrative engine that takes a
foundational text, decomposes it into a causal event graph, perturbs
load-bearing nodes via Pearl's do-calculus, and propagates coherent
counterfactual trajectories forward through the graph using the
Abduction–Action–Prediction algorithm. Unlike prior counterfactual
story-generation systems, which operate at depth one against a single
original, Rukha operates recursively to depths five and six: a
perturbation downstream of a previous perturbation produces a
counterfactual-of-a-counterfactual, and so on. Our central claim is that
recursive depth, properly scaffolded, is computationally tractable
where unscaffolded long-form generation collapses. Three convergences
make it so. First, formal: Pearl's framework, Lewis–Ryan possible-worlds
narratology, the Aaronson–Watrous \textsc{pspace} characterization of
multiverse computation, and the 5D Brainfuck \textsc{mtt} esolang all
describe the same structure --- a directed acyclic graph of branching
states governed by coherence constraints. Second, structural: persistent
data structures with structural sharing (hash array mapped tries,
Merkle-\textsc{dag}s, zero-suppressed decision diagrams) reduce the
$O(b^d)$ apparent state space to storage proportional to actual
divergence between universes. Third, architectural: explicit causal
\textsc{dag}s scaffold what unscaffolded chain-of-thought cannot ---
recent work documents reasoning-horizon decay at \mbox{70–85\%} of chain
length and near-random performance on nested counterfactuals without
external structure. We demonstrate the system on the Book of Jonah,
whose 16 consensus events and quantified chiastic symmetry --- the
39-Hebrew-word parallel between Jonah 4:2–3 and 4:10–11, machine-verifiable
against the Westminster Leningrad Codex --- make it the optimal minimal
test case. We close by arguing that Rukha is best understood as a
\emph{computational midrash} engine: a 2,000-year scholarly tradition of
generating textual counterfactuals as interpretive practice, made
recursive and explicit. The output is not a dataset of alternate
endings but a topology of grace.

\vspace{2.2em}
\diamondrule
\clearpage

% ─── Body ────────────────────────────────────────────────────────────
\setcounter{page}{1}
\maketitle
\vspace{-0.4em}

\begin{center}
\begin{minipage}{0.78\textwidth}\itshape\small\centering
The things that persist are the things that couple to their substrate
rather than consume it. Stories couple to their tellers, and to the
worlds they could have been.
\end{minipage}
\end{center}

\taperedrule
\vspace{0.6em}

\section{Introduction: A Story Is a Graph of Forks}

Current language models are trained on what humans wrote. This
limitation is rarely named but it is severe. The corpus of human
storytelling is the surviving trajectory of a much larger possibility
space --- every story we did not tell, every ending the author cut, every
counterfactual the redactor smoothed. A model trained only on the
surviving trajectory inherits the survivor bias. It learns what stories
\emph{are} but not what they \emph{could have been}, and consequently it
learns narrative \emph{causation} only as a shadow on the wall of
narrative \emph{outcome}.

Rukha is a proposal to lift the shadow. The system takes a foundational
text --- the Book of Jonah, the Mahābhārata, the Iliad, the Brothers
Karamazov --- decomposes it into a causal event graph using
constrained-decoding event extraction, identifies which nodes carry the
graph's structural load, perturbs those nodes through Pearl's
do-calculus, and propagates coherent alternative trajectories forward
via the Abduction–Action–Prediction (\textsc{aap}) algorithm. The output
is a corpus of causally-coherent alternate histories: not random
mutations, not GPT-fabulations, but lawful trajectories through the
narrative-physics established by the source material.

What distinguishes Rukha from prior work in counterfactual story
generation is depth. \textsc{TimeTravel} \cite{timetravel}, \textsc{educat}
\cite{educat}, the \textsc{pasta} dataset \cite{pasta}, and the broader
counterfactual-rewriting literature \cite{sketchcustomize} all operate at
depth one: a single ``what if'' applied to a single original story. Rukha
operates recursively. A perturbation downstream of a previous
perturbation produces a counterfactual-of-a-counterfactual; that, in
turn, can be perturbed again. We aim for depth five and six.

This is not a marginal extension of prior work. It is the regime where
counterfactual reasoning becomes interesting --- and the regime where
unscaffolded language-model generation has been shown to fail. Recent
empirical work documents a \emph{Reasoning Horizon} at roughly
\mbox{70–85\%} of chain length, beyond which intermediate reasoning
tokens have negative effect on the final output \cite{faithfulnessdecay}.
Other work documents \emph{Shallow Lock-in} and \emph{Deep Decay} as
characteristic failure modes of long-chain reasoning
\cite{stepflowbreaks}. \textsc{counterbench} \cite{counterbench}, a 2025
benchmark for counterfactual reasoning in language models, finds that
leading models perform near-randomly on its \emph{Nested} subset --- the
condition closest to Rukha's regime. Rukha's bet is that this collapse is
not fundamental; it is a consequence of asking models to perform
recursive counterfactual reasoning \emph{without external scaffolding}.
Given an explicit causal graph, persistent state, and the
Abduction–Action–Prediction protocol enforced by constrained decoding,
the bet is that recursive depth becomes tractable.

\subsection{What This Paper Argues}

The paper makes four claims and defends each.

\begin{enumerate}
  \item \emph{Formal equivalence.} Pearl's structural causal models,
    David Lewis's possible-worlds semantics, Marie-Laure Ryan's textual
    actual world plus modal satellite worlds, and the multiverse-time-travel
    primitives of the 5D Brainfuck \textsc{mtt} esolang all describe the
    same underlying structure: a directed acyclic graph of branching
    states governed by coherence constraints. Rukha is the operationalization
    of this structure.

  \item \emph{Complexity tractability.} Recursive counterfactual
    propagation on causal event graphs lives within \textsc{pspace}, by
    Aaronson–Watrous \cite{aaronsonwatrous}. Persistent data structures
    with structural sharing --- hash array mapped tries
    \cite{bagwellhamt}, Merkle-\textsc{dag}s as in Git and \textsc{ipfs},
    and zero-suppressed decision diagrams \cite{knuthzdd} --- reduce
    apparent $O(b^d)$ storage to storage proportional to the actual
    divergence between universes.

  \item \emph{Recursion depth is the empirical frontier.} No published
    system has demonstrated coherent counterfactual reasoning at depth
    five or six on natural-language narrative. The literature has
    \emph{named} the problem (\textsc{counterbench-Nested}, the Reasoning
    Horizon, Shallow Lock-in / Deep Decay) but not solved it. This is
    Rukha's contribution.

  \item \emph{Sacred narrative is the optimal demonstration target.}
    The Book of Jonah, with its 48 verses, 16 consensus events, and
    quantified chiastic structure --- the 39-Hebrew-word parallel between
    Jonah 4:2–3 and 4:10–11, machine-verifiable against the Westminster
    Leningrad Codex --- is the smallest text where Rukha's claims can be
    rigorously demonstrated. The theological theme of the book ---
    the disorientation of receiving mercy you did not think the
    recipient deserved --- becomes computational under perturbation.
\end{enumerate}

\diamondrule

\section{Foundations}

\subsection{Pearl's Do-Calculus and the AAP Algorithm}

A structural causal model (\textsc{scm}) over variables $V = \{V_1,
\dots, V_n\}$ specifies, for each $V_i$, a structural equation $V_i =
f_i(\text{pa}(V_i), U_i)$ where $\text{pa}(V_i)$ are the causal parents
of $V_i$ and $U_i$ is exogenous noise. The do-operator $\text{do}(V_i :=
v')$ surgically replaces $f_i$ with the constant assignment $V_i = v'$,
severing all incoming edges to $V_i$ while leaving downstream structural
equations intact. This is Pearl's central technical move
\cite{pearlcausality}: an intervention is structural surgery, not
conditioning.

Counterfactual reasoning in this framework follows the
Abduction–Action–Prediction algorithm \cite{pearlalgorithmization,
pearlrobins}:

\begin{enumerate}
  \item \sceq{Abduction.} Given the observed facts $E$ (the evidence
    encoded in the actual textual world), update the prior over
    exogenous noise: $P(U) \mapsto P(U \mid E)$. In narrative terms,
    abduction infers the latent assumptions about characters, settings,
    and world-rules that are compatible with what we read.
  \item \sceq{Action.} Apply the intervention: $\text{do}(V_i := v')$.
    The structural equation for $V_i$ is replaced; incoming edges are
    severed.
  \item \sceq{Prediction.} Forward-propagate the modified \textsc{scm}
    in topological order. For each downstream node $V_j$, compute
    $V_j = f_j(\text{pa}(V_j), U_j)$ using the updated $P(U \mid E)$ and
    the modified parent values.
\end{enumerate}

In Rukha, each step is performed by a 70B-parameter language model under
\textsc{xgrammar}-constrained \textsc{json} decoding
\cite{xgrammar}. The structural equations $f_j$ are not closed-form
functions; they are prompts. The model is asked, given the observed
narrative and a perturbation at $V_i$, to predict the new value of $V_j$
in a manner consistent with the perturbed graph and the abducted
exogenous priors. The grammar constrains the output schema --- event
\textsc{id}, type, agents, location, verse-reference --- so that
hallucination of structure is impossible at the syntactic level. What
remains is the semantic problem: did the model abduct correctly, did it
propagate correctly, did it preserve coherence?

\subsection{Recursion: The Pearl–Robins Decomposition}

Sequential interventions in acyclic \textsc{scm}s decompose cleanly.
This is the foundational result for nested counterfactuals: Pearl and
Robins \cite{pearlrobins} establish that compound interventions in
\textsc{dag}s with hidden variables can be evaluated as a chain of
single-step do-operations, each conditioned on the consequences of those
preceding it. Correa, Lee, and Bareinboim \cite{correaleebareinboim}
extend this to nested counterfactual identification from arbitrary
surrogate experiments, providing a sound and complete algorithm for the
identifiability question.

The implication: a depth-$N$ recursive counterfactual is mathematically
well-defined whenever the underlying graph is acyclic and the
identification conditions are satisfied. Rukha's narrative graphs are
acyclic by construction --- causal edges flow forward in narrative time
--- and the identification conditions are trivially satisfied because
all events are observable. There is no hidden-variable confounding in a
fixed text: the text is the data.

\begin{proposition}[Recursive AAP Decomposition]
Let $G = (V, E)$ be an acyclic narrative event graph and let
$(v_{i_1}, v'_{i_1}), \dots, (v_{i_N}, v'_{i_N})$ be a sequence of
perturbations such that $v_{i_k}$ is reachable from $v_{i_{k-1}}$ in $G$
for all $k > 1$. Then the depth-$N$ recursive counterfactual
$G^{(N)}$ is the result of applying \textsc{aap} sequentially:
$G^{(0)} = G$;
$G^{(k)} = \mathrm{AAP}(G^{(k-1)}, v_{i_k} \mapsto v'_{i_k})$.
The decomposition is well-defined and unique up to the abduction
distribution at each step.
\end{proposition}

What this proposition does \emph{not} guarantee is that the
language-model implementation of \textsc{aap} is faithful to the
abduction distribution at each step. It does not guarantee that depth
five or six is computationally achievable in practice. These are the
empirical questions the paper later confronts.

\subsection{Possible-Worlds Narratology}

Rukha's perturbations are not statistical; they are semantic. The
language for talking about narrative possibility is
narratology, specifically Marie-Laure Ryan's possible-worlds theory
\cite{ryan}.

Ryan distinguishes the \emph{textual actual world} (\textsc{taw}) ---
what is presented as factual within the fiction --- from the
\emph{satellite worlds} of characters: their \emph{knowledge worlds}
(what they believe), \emph{wish worlds} (what they want), \emph{obligation
worlds} (what they should do), \emph{intention worlds} (what they plan
to do). Each character is a center of multiple modal worlds, and
narrative tension arises when the satellite worlds of one character
conflict with the \textsc{taw} or with the satellite worlds of another.

Lewis \cite{lewiscounterfactuals} grounds counterfactual reasoning in
similarity orderings over possible worlds: the counterfactual ``if $A$
had not happened, $B$ would not have happened'' is true if and only if
in the closest possible world where $A$ does not happen, $B$ does not
either. Ryan adapts Lewis's framework to fiction via the
\emph{principle of minimal departure}: readers reconstrue the fictional
world as the closest possible to the actual world, deviating only where
the text forces deviation.

These two frameworks --- Pearl's \textsc{scm} and Ryan-Lewis modal
worlds --- describe the same object from two angles. Pearl asks: what
happens if I surgically modify this node? Lewis asks: what is the closest
world where the modification holds? In an acyclic narrative \textsc{dag}
with no hidden confounders, the answers coincide: the
\textsc{aap}-propagated counterfactual \emph{is} the closest possible
world.

\subsection{Complexity: Why \textsc{pspace}?}

The ceiling of Rukha's computational regime is established by
Aaronson and Watrous \cite{aaronsonwatrous}: classical and quantum
computation in the presence of closed timelike curves under Deutsch's
self-consistency condition is exactly \textsc{pspace}-complete. Their
result formalizes a long folkloric intuition: time-travel computation,
properly constrained, is no more powerful than polynomial space.

Rukha is not implementing closed timelike curves. Its counterfactual
trajectories propagate forward; they do not loop back. But the ambient
class within which Rukha operates --- bounded-depth multiverse exploration
with consistency constraints --- is a strict subset of
\textsc{pspace}, and Aaronson and Watrous's result anchors the
upper bound.

The 5D Brainfuck With Multiverse Time Travel esolang
\cite{fivedbfmtt} provides an engineering-folkloric instantiation of the
same idea: a programming language whose primitives include spawning
parallel timelines, killing pointers across timelines, and rewinding
within timelines. The esolang is, deliberately, a minimal model of
\textsc{pspace}-class multiverse computation. Rukha can be read as the
narrative-domain instantiation of the same primitives:
\texttt{spawn-timeline} = perturbation, \texttt{kill-pointer} = quality
filtering, \texttt{rewind} = re-abduction.

\diamondrule

\section{The Recursion-Depth Claim}

\subsection{Why Depth Matters}

Counterfactual reasoning at depth one is well-trodden territory.
\textsc{TimeTravel} \cite{timetravel} provides 29,849 supervised single-step
counterfactual rewrites of \textsc{rocstories} narratives;
\textsc{educat} \cite{educat} achieves unsupervised single-step rewriting
via causal-effect estimation. The \textsc{pasta} dataset \cite{pasta}
introduces participant-state perturbations as a cleaner formalism. None
of these systems propagates recursively. Each ``what if'' is applied to
the canonical original; counterfactuals do not stack.

The reason this matters is that depth is where narrative \emph{physics}
becomes visible. At depth one, a perturbation reveals the immediate
causal neighbors of an event. At depth three, perturbations reveal which
\emph{patterns} survive when the structural scaffolding is repeatedly
disturbed. At depth five and six, an empirical question becomes
testable: do generated counterfactual variants cluster into
narrative \emph{archetypes}, or do they spread uniformly through
possibility space? The answer to this question is the difference between
narrative being a soft topology with attractor basins (Propp's
functions, Campbell's monomyth, Hogan's prototypical structures, Reagan's
six emotional arcs \cite{reagansixarcs}) and narrative being an
arbitrary high-dimensional space.

If narrative \emph{has} attractor basins, then deep counterfactual
exploration should reveal them: most depth-six variants of Jonah should
land in a small number of recurrent shapes, regardless of which specific
nodes were perturbed. If narrative \emph{lacks} attractor basins, then
depth-six variants should fan out and not return. Either result is
publishable. The first vindicates Propp, Campbell, and Hogan as
empirical claims about narrative cognition; the second falsifies them.

\subsection{The Empirical Landscape}

\subsubsection{What Has Been Shown}

The most directly relevant 2026 result is Mechanistic Faithfulness
Decay in Chain-of-Thought Reasoning \cite{faithfulnessdecay}. The authors
introduce Normalized Logit Difference Decay (\textsc{nldd}) and identify
a \emph{Reasoning Horizon} $k^*$ at \mbox{70–85\%} of chain length, beyond
which reasoning tokens have ``little or negative effect'' on the final
answer. For unscaffolded reasoning chains, this implies an empirical
ceiling on coherent recursion depth.

\textsc{counterbench} \cite{counterbench} extends this to counterfactual
reasoning specifically. The benchmark contains 1,200 questions across
five \textsc{scm} types --- Basic, Joint, \emph{Nested}, Conditional,
Backdoor. The result for the Nested condition is striking: leading
language models perform \emph{near-randomly} without external structural
scaffolding, and the authors' proposed CoIn algorithm --- explicit
search-and-backtrack over the \textsc{scm} --- substantially closes the
gap. This is direct empirical support for the hypothesis Rukha rests on:
nested counterfactuals are within reach when the structural scaffolding
is explicit, and out of reach when it is not.

Vashishtha et al.'s \emph{Executable Counterfactuals}
\cite{executablecounterfactuals}, accepted at \textsc{iclr} 2026,
decomposes counterfactual reasoning into Pearl's three steps and reports
that prior benchmarks routinely \emph{skip the abduction step} and
consequently overestimate language-model performance. The authors find
that supervised fine-tuning generalizes poorly across counterfactual
\textsc{scm}s and propose code-as-\textsc{scm} prompting. The
implication for Rukha is a methodological imperative: the
\textsc{aap} protocol must be enforced \emph{at each recursion step},
not assumed, and the abduction step in particular must be made explicit
in the prompt structure.

\textsc{step Flow Breaks} \cite{stepflowbreaks} documents two failure
modes that systematically distinguish correct from incorrect long-chain
reasoning traces: \emph{Shallow Lock-in}, in which the model commits to
an incorrect partial trajectory and cannot recover, and \emph{Deep Decay},
in which the model loses track of structural constraints over many
steps. The authors propose a test-time intervention they call
\textsc{stepflow}; the relevance to Rukha is that recursive
counterfactual propagation is exactly the regime where these failure
modes are expected to dominate, and Rukha's structural-sharing
infrastructure is, by design, an anti-Shallow-Lock-in mechanism (the
graph state is persistent across recursion levels) and the
\textsc{aap} protocol is, by design, an anti-Deep-Decay mechanism (each
step is grounded in the abducted prior plus the graph state).

\subsubsection{What Has Not Been Shown}

To our knowledge, no published system has demonstrated coherent
counterfactual reasoning at depth $\geq 3$ on natural-language narrative.
\textsc{TimeTravel}, \textsc{educat}, \textsc{pasta}, and the broader
counterfactual story-generation literature operate at depth one.
\textsc{counterbench} addresses formal-\textsc{scm} counterfactual
question-answering, not narrative generation. The recent depth-generalization
literature in symbolic reasoning \cite{depthgeneralization} works on
Boolean algebra, recursive arithmetic, and propositional logic; not
narrative. \textsc{deepImagine} \cite{deepimagine}, the closest semantic
match for ``successive counterfactual imagining,'' targets biomedical
reasoning rather than narrative and does not foreground depth.

The empirical frontier Rukha targets is \emph{exactly} the gap between
the depth-one counterfactual story-generation literature and the
formal-\textsc{scm} nested-counterfactual literature. Closing this gap
is the contribution.

\subsection{Why Scaffolding Should Help}

The architectural bet is that explicit causal-graph scaffolding plus
\textsc{aap}-protocol prompting plus structural-sharing memoization
jointly preserve coherence at depth five and six where unscaffolded
chain-of-thought collapses. The argument has three legs.

First, the model is not asked to perform unbounded long-form
generation. Each recursion step produces a small number of structured
event records, conformant to a \textsc{json} grammar, conditioned on the
abducted prior and the perturbed graph. The total token budget per step
is bounded; the Reasoning Horizon at \mbox{70–85\%} of chain length never
becomes the binding constraint.

Second, the structural-sharing infrastructure ensures that the
graph-state at each recursion level is persistent and content-addressable.
Identical sub-graphs at different levels resolve to the same Merkle
\textsc{cid}; updates to a sub-graph are localized; the model's view of
``the world up to this point'' is always exact. There is no accumulating
drift from compressed summaries.

Third, the \textsc{aap} protocol enforces that abduction is performed
\emph{explicitly} at each step. The model is asked, before propagating,
to articulate which exogenous priors are being preserved and which are
being updated. This is the methodological correction Vashishtha et al.
\cite{executablecounterfactuals} call for. Coupled with constrained
decoding, the protocol cannot be skipped.

Whether this scaffolding suffices is the empirical question of Section
6. We do not claim it does. We claim only that the question is testable,
that the setup is principled, and that no other published system is in a
position to test it.

\diamondrule

\section{Architecture}

\subsection{Pipeline Overview}

The Rukha pipeline operates in six sequential stages, each producing
artifacts that feed the next:

\begin{center}\small
\begin{tabular}{@{}c l p{3.4in}@{}}
  \toprule
  \sceq{Stage} & \sceq{Name} & \sceq{Operation} \\
  \midrule
  1 & Text Ingestion       & Load source text; segment by chapter / verse; assign metadata. \\
  2 & Event Extraction     & Constrained-decoding extraction of 14--18 event nodes per chapter. \\
  3 & Edge Identification  & Causal / temporal / conditional edge typing; strength score; evidence span. \\
  4 & DAG Construction     & NetworkX \texttt{DiGraph}; load-bearing scoring; acyclicity check. \\
  5 & Perturbation         & Select a load-bearing node; generate counterfactual alternatives. \\
  6 & Propagation          & \textsc{aap} forward-propagation in topological order to depth $N$. \\
  \bottomrule
\end{tabular}
\end{center}

Stages 5 and 6 are recursive. The propagated graph $G^{(1)}$ is treated
as the new \textsc{taw} and re-perturbed at depth 2; this iterates to
target depth $N \in \{5, 6\}$.

\subsection{Event Graph Extraction}

Extracting causal event graphs from narrative prose is the most
technically novel phase of the pipeline. No existing system targets
literary, mythological, or sacred-text event extraction with the
specificity Rukha requires. We adopt the \emph{code-as-schema} prompting
pattern validated by GoLLIE at \textsc{iclr} 2024 \cite{gollie} and
PyCode-TextEE at \textsc{acl} 2025 \cite{pycodetexteed}: schemas are
expressed as typed Python dataclasses with docstring guidelines, and the
model emits a structured \texttt{result = [\dots]} list. The empirical
finding from these two papers is that code-as-schema prompting forces
the model to follow the explicit definitions rather than fall back on
priors --- exactly the behavior Rukha needs.

Li, Pan, and Pi at the \textsc{acl} 2025 Workshop on Narrative
Understanding \cite{lihybrid} demonstrated that a hybrid linguistic
pipeline (RoBERTa + Expert Index + \emph{stac} classification) outperforms
GPT-4o and Claude 3.5 on narrative causal-graph quality across 100
chapters and short stories. The lesson, taken seriously, is that pure
language-model extraction is suboptimal for narrative causal graphs;
Rukha's extraction layer therefore combines constrained-decoding language-model
event extraction (Stages 2–3) with a deterministic
graph-validation pass (Stage 4) that prunes weakly-supported edges and
enforces acyclicity.

The Pydantic schema for an event is given in Appendix A. The key
fields are: \texttt{id}, \texttt{trigger}, \texttt{summary},
\texttt{event\_type} (one of \textsc{action}, \textsc{state\_change},
\textsc{decision}, \textsc{divine\_action}, \textsc{communication}),
\texttt{stac\_label} (Situation, Task, Action, Consequence),
\texttt{agents}, \texttt{location}, \texttt{verse\_ref}.

\subsection{Load-Bearing Scores}

Not every event is structurally critical. The chiastic structure of
Jonah does not depend on whether the storm in chapter 1 is described as
``mighty'' or ``great''; it depends on whether God commissioned Jonah at
the beginning. Identifying which nodes carry the graph's structural
load is the perturbation engine's first task.

We compute a composite load-bearing score:
\[
  \text{load}(v) = 0.40 \cdot \text{BC}(v) + 0.30 \cdot \text{PR}(v) + 0.30 \cdot \text{OD}(v)
\]
where $\text{BC}(v)$ is Freeman's betweenness centrality
\cite{freemanbetweenness}, computed via Brandes's $O(VE)$ algorithm;
$\text{PR}(v)$ is PageRank \cite{pagerank}; and $\text{OD}(v)$ is
out-degree, normalized to the unit interval. Events scoring above the
80th percentile are classified as \emph{Critical}; between the 40th and
80th, \emph{Significant}; below 40th, \emph{Decorative}. The
perturbation engine targets Critical and Significant nodes.

These weights are heuristic. A real ablation study --- correlating
$\text{load}(v)$ with empirical post-\textsc{aap} graph-edit distance ---
is part of the evaluation protocol described in Section 6. Halpern–Pearl
actual causality \cite{halpernpearlcausality} provides the principled
alternative: $v$ is load-bearing if and only if changing $v$ via
$\text{do}(\cdot)$ changes a substantial fraction of downstream events
under counterfactual evaluation. The composite score is a topological
proxy for this; the proxy is testable.

\subsection{Perturbation Strategies}

The perturbation engine generates plausible counterfactual alternatives
at each load-bearing node. Six strategies, each targeting a distinct
narrative operation:

\begin{center}\small
\begin{tabular}{@{}l p{2.4in} l@{}}
  \toprule
  \sceq{Strategy} & \sceq{Description} & \sceq{Example (Jonah E2)} \\
  \midrule
  \textsc{inversion}        & Event replaced by its opposite           & Jonah obeys immediately \\
  \textsc{replacement}      & Event replaced by a different action     & Jonah flees to Egypt instead \\
  \textsc{removal}          & Event deleted                            & No flight; Jonah deliberates \\
  \textsc{intensification}  & Event amplified                          & Jonah leads armed rebellion \\
  \textsc{temporal\_shift}  & Event occurs at a different time         & Jonah flees \emph{after} seeing Nineveh \\
  \textsc{transplant}       & Event grafted from another text          & Jonah's crisis mirrors Arjuna's \\
  \bottomrule
\end{tabular}
\end{center}

Each perturbation must satisfy four coherence constraints: world
consistency (the alternative must be physically and logically possible
within the established world-rules), character consistency (characters
must act compatibly with their established traits, unless the
perturbation explicitly targets character), causal validity (no
contradictions with non-perturbed upstream events), and
counterfactual invariance (events $d$-separated from the perturbation
remain unchanged --- the \textsc{TimeTravel} principle, that on average
1.5 of 5 sentences change \cite{timetravel}).

\subsection{Structural Sharing: Why Depth-Six Is Tractable}

Naive depth-$N$ counterfactual exploration branches combinatorially.
With branching factor $b = 5$ alternatives per critical node, depth 5
yields $5^5 = 3{,}125$ universes; depth 6 yields $15{,}625$. Each
universe is itself a 16-event narrative graph. The naive storage
requirement is intractable.

The argument that it is in fact tractable rests on a structural
observation: sibling counterfactual universes share most of their
sub-structure. Most events do not change downstream of most
perturbations. The actual divergence between any two universes is
small, and the storage requirement scales with divergence, not with
universe count.

The mechanism is borrowed from the persistent-data-structures literature
\cite{driscolltarjan, okasakipfds}. Three layers compose:

\begin{itemize}
  \item \emph{Hash consing} \cite{filliatreconchon}: structurally
    identical objects share a single representation. Equality testing
    becomes pointer comparison ($O(1)$); de-duplication is global.
  \item \emph{Hash array mapped tries} \cite{bagwellhamt}: 32-way
    branching with bitmap+popcount compression; expected depth
    $\log_{32}(n)$. Each universe's per-character or per-event state is
    a \textsc{hamt}; updates copy approximately four nodes while sharing
    the remaining structure.
  \item \emph{Merkle-\textsc{dag}s}: as deployed in Git and
    \textsc{ipfs} \cite{ipfsmerkledag}. Each graph node is keyed by the
    hash of its content (\texttt{cid}). Forking a universe is $O(1)$;
    structural-equality between two sub-graphs is $O(1)$; locating
    the differences between two universes is $O(\log n)$.
  \item \emph{Zero-suppressed decision diagrams} \cite{knuthzdd},
    Knuth's \textsc{zdd}, for cross-universe set queries. The famous
    benchmark: 437 billion paths from California to Maine on the US
    capitals graph compress into 7,850 \textsc{zdd} nodes --- a
    ratio of $5.6 \times 10^7$.
\end{itemize}

In aggregate: storage scales not with the apparent
$O(b^d)$ but with the actual edit distance between universes, which
empirical observation suggests is small. We will report the achieved
compression ratio in the evaluation.

\diamondrule

\section{The Jonah Demonstration}

\subsection{Why Jonah}

The Book of Jonah is the optimal proof-of-concept target. It has 48
verses, approximately 1{,}273 Hebrew words, and decomposes into 16
consensus load-bearing events. It is self-contained, narratively dense,
and possesses a quantified chiastic structure that gives Rukha a hard
ground-truth target: any extraction algorithm worth its weight should
discover the chiasm.

\subsection{The Sixteen Events}

\begin{center}\small
\begin{tabular}{@{}c c p{2.6in} l l@{}}
  \toprule
  \sceq{ID} & \sceq{Verse} & \sceq{Event} & \sceq{Type} & \sceq{Weight} \\
  \midrule
  E1  & 1:1–2  & God commissions Jonah to preach against Nineveh & \textsc{divine}  & Critical \\
  E2  & 1:3    & Jonah flees to Tarshish                         & \textsc{decision}  & Critical \\
  E3  & 1:4    & God sends a great storm                         & \textsc{divine}    & Significant \\
  E4  & 1:7    & Lots cast; Jonah identified                     & \textsc{action}    & Significant \\
  E5  & 1:15   & Jonah thrown overboard                          & \textsc{action}    & Significant \\
  E6  & 1:17   & Great fish swallows Jonah                       & \textsc{divine}    & Significant \\
  E7  & 2:1–9  & Jonah's prayer from the deep                    & \textsc{communication} & Decorative \\
  E8  & 2:10   & Fish vomits Jonah onto land                     & \textsc{divine}    & Significant \\
  \midrule
  E9  & 3:1–2  & God's second commission                         & \textsc{divine}    & Critical \\
  E10 & 3:4    & Jonah preaches in Nineveh                       & \textsc{action}    & Significant \\
  E11 & 3:5–9  & Nineveh repents                                  & \textsc{state\_change} & Critical \\
  E12 & 3:10   & God relents from destruction                    & \textsc{divine}    & Critical \\
  E13 & 4:1–3  & Jonah's anger at God's mercy                    & \textsc{state\_change} & Significant \\
  E14 & 4:6    & God appoints a plant for shade                  & \textsc{divine}    & Decorative \\
  E15 & 4:7    & Worm destroys the plant                         & \textsc{divine}    & Decorative \\
  E16 & 4:10–11 & God's final rhetorical question                & \textsc{communication} & Critical \\
  \bottomrule
\end{tabular}
\end{center}

\subsection{The Chiastic Structure}

The book exhibits a macro-chiastic mirror structure. Movement 1
(chapters 1–2: Jonah and the sailors and the fish) parallels Movement 2
(chapters 3–4: Jonah and the Ninevites and the plant): E1 mirrors E9
(two commissions), E2 mirrors E10 (flight versus obedience), E5 mirrors
E11 (thrown to sea versus thrown to repentance), E6 mirrors E12 (fish
saves Jonah versus God saves Nineveh), E7 mirrors E13 (prayer versus
anger). The structure is recognized across major commentaries
\cite{trible, sasson, magonet}.

The Hebrew text quantifies the symmetry. Magonet \cite{magonet}
demonstrates that Jonah 4:2–3 (Jonah's complaint after Nineveh repents)
contains exactly 39 Hebrew words, and Jonah 4:10–11 (God's reply about
the qiqayon plant) contains exactly 39 Hebrew words. Counting maqqef-joined
clusters as single units yields 30 in both passages. The dual identity
across both counting conventions is what makes the parallel structurally
striking and what indicates intentional artistry rather than accident
\cite{magonettorah}.

This count is machine-verifiable. The Open Scriptures Hebrew Bible
project \cite{oshb} provides the Westminster Leningrad Codex in OSIS
\textsc{xml} with each \texttt{<w>} element marking a graphic word.
Iterating over the verses 4:2–3 and 4:10–11 of Jonah and counting
\texttt{<w>} elements --- with maqqef-joined sequences either treated as
two words or collapsed --- reproduces Magonet's figures exactly.
The verification script and methodology are given in Appendix B; the
script-derived count is what we report in the paper, with Magonet cited as
the originating literary observation.

\subsection{The Demonstration Perturbation}

The most theologically charged depth-one perturbation is the
\textsc{inversion} at E1: \emph{What if Jonah obeyed at the first
commission?}

The naive expectation is that obedience eliminates the entire sea
adventure. Events E2 through E8 collapse: no flight, no storm, no
sailors, no fish, no prayer from the deep. The story jumps directly
from E1 to E10 (Jonah preaching in Nineveh).

But here Jonah 4:2 reveals the theological irony that makes the demo
non-trivial. After Nineveh repents and God relents, Jonah complains:
``\emph{This is what I said when I was still in my own country; that
is why I fled to Tarshish at the beginning. For I knew that you are a
gracious God and merciful, slow to anger, and abounding in steadfast
love, and ready to relent from punishing.}'' Jonah did not flee from
prophetic failure; he fled \emph{because he knew God would show mercy}
and he could not bear it. The expected counterfactual is exactly what
happens canonically, just without the detour through the sea.

The depth-one perturbation therefore poses the test question: \emph{does
Jonah still rage at God's mercy in the world where he did not suffer
first?}

The depth-five and depth-six perturbations recurse on this question.
\textsc{do(E1, obeys-immediately)} produces $G^{(1)}$. Within $G^{(1)}$,
perturbing E11 (\textsc{do(E11, Nineveh-does-not-repent)}) produces
$G^{(2)}$: Jonah preaches as commanded but the Ninevites refuse.
Within $G^{(2)}$, perturbing the implicit divine-response node
(\textsc{do(E12, God-destroys-Nineveh-anyway)}) produces $G^{(3)}$.
At each recursion level the model must abduct the new exogenous priors,
sever the appropriate causal edges, and propagate. By depth five we are
asking about a world in which Jonah obeyed, Nineveh refused, God
destroyed, the survivors fled to Tarshish, the survivors' prophet
returned, and so on. The empirical question is whether the depth-five
narrative remains coherent --- in particular, whether Jonah's
psychological architecture (the rage at mercy he did not extend himself)
remains recognizable.

\diamondrule

\section{Evaluation Methodology}

A claim about depth-five recursive coherence is empirically vacuous
without a measurement protocol. We propose triangulation across four
methodological registers.

\subsection{Per-Depth Consistency Error Density}

The first register is automatic: ConStory-Bench \cite{constory}
introduces Consistency Error Density (\textsc{ced}) as the canonical
metric for long-form narrative consistency, and ConStory-Checker
operationalizes it as an \textsc{llm}-judge protocol with grounded
evidence quotation across a 19-subtype error taxonomy. We adopt
\textsc{ced} but compute it \emph{at each recursion level
independently}: $\text{CED}(G^{(k)})$ for $k \in \{0, 1, 2, 3, 4, 5, 6\}$.

The hypothesis Rukha must defend is that $\text{CED}$ does \emph{not}
grow super-linearly with $k$. A linear or sub-linear curve, even with a
non-trivial intercept, vindicates the structural-sharing thesis.
Super-linear growth --- particularly an inflection between depth 3 and 4
--- would indicate that the \textsc{aap} protocol is not in fact
preserving abduction across recursion levels and the depth claim
collapses.

\subsection{Minimal-Pair Discrimination}

The second register is comparative: the NoCha protocol
\cite{nocha} evaluates whether a model can distinguish minimally-different
True/False claims about a long narrative. We adapt the protocol: for
each generated branch $G^{(k)}$, we generate a sibling branch $G^{(k)}_{\text{sib}}$
by perturbing a different node at the same recursion level, and ask the
model to discriminate factual claims about the two. A successful
demonstration of Rukha at depth $k$ requires that the model's
discrimination accuracy on $(G^{(k)}, G^{(k)}_{\text{sib}})$ pairs be
substantially above chance.

Discrimination accuracy at depth $k$ is therefore the most direct test
of whether ``the depth-$k$ counterfactuals are actually different
stories'' or ``the depth-$k$ counterfactuals are the same story with
labels swapped.''

\subsection{Character-State Knowledge-Graph Consistency}

The third register is structural: NarrativeFactScore \cite{narrativefactscore}
introduces the Character Knowledge Graph as an Agent-as-Judge oracle for
narrative factuality. We extend the methodology: for each generated
branch, a separate model maintains a per-character state graph
(beliefs, locations, possessions, intentions) updated incrementally as
events unfold. Consistency violations --- a character believing $X$ at
event $E_i$ and $\neg X$ at event $E_{i+1}$ without a state-change
event --- are flagged and counted.

This is the symbolic-state-tracking approach validated by SCORE
\cite{scorestory}, which reports a 23.6\% improvement in coherence and a
41.8\% reduction in hallucinations against unscaffolded baselines.
Rukha's bet is that depth-five recursive generation \emph{requires}
this scaffolding; we report the per-depth consistency-violation rate as
a function of recursion level.

\subsection{Human Evaluation}

The fourth register is human. Automatic metrics --- even agent-as-judge
ones --- have known correlation gaps with human judgment on long-form
narrative. We adopt a pairwise-preference protocol at depths 3, 5, and 6:
human evaluators are presented with $(G^{(k)}, G^{(k)}_{\text{sib}})$
pairs and asked to assess which branch is (a) more internally coherent,
(b) more recognizably a continuation of Jonah, and (c) more
interpretively interesting. The third question is the most important
and the least quantifiable; it is also the question Rukha is actually
trying to answer.

\subsection{The Cultural-Attractor Diagnostic}

A separate evaluation strand probes the narrative-attractor hypothesis.
Perez et al. \cite{perezattractors} demonstrate that iterated
language-model transmission converges to attractor states for
toxicity, positivity, difficulty, and length. We adapt their
methodology to ask: do Rukha's depth-six counterfactual variants of
Jonah cluster into archetypal shapes (Reagan's six emotional arcs;
Hogan's heroic / romantic / sacrificial prototypes; Propp's morphological
functions \cite{proppmorphology}), or do they distribute uniformly?

Operationally, we embed each depth-six variant via sentence-transformers,
project into the six-arc emotional-trajectory space following Reagan et
al. \cite{reagansixarcs}, and cluster. If clusters form along
archetypal lines, narrative attractors are real and computationally
visible. If they do not, the attractor hypothesis fails empirically.

This is the single most novel empirical contribution Rukha makes
beyond the depth claim itself: a quantitative test of whether narrative
archetypes are basins of attraction in counterfactual space. The
narratological tradition (Propp, Campbell, Hogan) has asserted this for
a century; Rukha is the first system in a position to falsify it.

\diamondrule

\section{What Rukha Is For}

\subsection{The Immediate Output}

The immediate output of a Rukha campaign is a structured corpus.
Each entry is a complete causally-coherent alternate trajectory
through a foundational text, indexed by its perturbation path,
content-addressed by Merkle \textsc{cid}, and quality-scored by the
evaluation pipeline.

The corpus has obvious applications. Training data for models that need
to understand causal narrative reasoning --- not just predict the next
word but model the consequences of structural change to a fixed
narrative. Comparative-mythology research at scales no single scholar
could undertake: every permutation of every perturbation of every
foundational text in the seed corpus, queryable by archetype and by
edit-distance from canonical. Psychological research on counterfactual
thinking with controlled stimuli that vary in exactly the dimensions a
researcher cares about. A living, extensible dataset that grows as new
texts are added and new perturbation strategies are validated.

\subsection{The Mediated Output: Narrative Physics}

A more interesting output is the empirical map of narrative possibility
that Rukha makes visible. Stories are not arbitrary high-dimensional
walks. Patterns recur: the descent and ascent, the prodigal return, the
sacrifice that fails, the mercy that wounds the merciful. Whether these
patterns are universal cognitive features (Hogan), Jungian archetypes
(Campbell), folkloristic functions (Propp), or simply emotional-arc
attractors (Reagan) has been a century-long humanistic question
without empirical traction. Rukha provides the traction.

\begin{claim}[The Narrative-Physics Hypothesis]
Under recursive counterfactual perturbation of foundational texts at
depth $\geq 4$, the resulting variant trajectories cluster into a small
number of recurrent shapes with measurable basins of attraction in
emotional-arc space.
\end{claim}

The claim is testable. Rukha is the test apparatus. If the claim is
true, an entire branch of narratology becomes computational and
quantitative. If it is false, an entire branch of narratology has been
falsified.

\subsection{The Reflexive Output: What Counterfactuals Reveal}

The deepest output is interpretive. A counterfactual is not a
hypothetical; it is a tool for reading. The question
\emph{``what if Jonah obeyed at chapter 1?''} produces a story about
Nineveh. But the question \emph{``what if Jonah had been unable to flee
in the first place?''} produces a story about Jonah's psychology. And
the recursive question --- \emph{what if Jonah obeyed, Nineveh repented,
Jonah raged anyway, God responded with the same rhetorical question
about mercy} --- produces a story about how stubborn human resistance to
grace is, even when stripped of its operational excuses.

Each perturbation reveals what the canonical text was actually about.
Depth-one tells you what the events depend on; depth-three tells you
what the characters depend on; depth-five tells you what the
\emph{theology} depends on. This is the same operation that midrash
performs: by asking what \emph{could have been}, the tradition surfaces
what \emph{is}.

\diamondrule

\section{Computational Midrash: Ethics and Framing}

A system that generates counterfactual versions of sacred texts will
generate strong reactions. We do not pretend otherwise. We argue that
the framing matters, and we propose three.

\subsection{The Midrashic Tradition}

Rabbinic midrash is a 2,000-year scholarly tradition of generating
textual counterfactuals as interpretive practice. Aggadic midrash in
particular routinely asks \emph{what if} questions of biblical
narrative: what was Jonah doing in the fish's belly; what did Sarah
think when Abraham took Isaac up the mountain; what did the angels say
when humans were created. These questions are not transgressions; they
are the textual tradition's standard interpretive method. \emph{Pirkei
Avot} 5:22 instructs the reader of Torah to \emph{``turn it and turn it,
for everything is in it''} --- a 2,000-year-old endorsement of recursive
re-interpretation.

Rukha is computational midrash. It is the same operation, made
algorithmic and made recursive. The objection that ``you should not
generate counterfactual versions of sacred texts'' has an answer: this
is what the tradition has always done. The novelty is in the tooling,
not the question.

\subsection{Counterfactual History as Methodological Precedent}

A second framing comes from outside theology. Niall Ferguson's
\emph{Virtual History} \cite{virtualhistory} establishes counterfactual
history as a methodologically rigorous practice in the humanities. The
constraint Ferguson imposes --- that the counterfactual must be one
\emph{contemporaries actually considered} --- is the right constraint for
Rukha to engage. We do not perturb arbitrarily; we perturb at
nodes where the text itself foregrounds alternative possibilities. Jonah
4:2 explicitly states that Jonah \emph{contemplated} obedience and chose
flight. The Ninevites \emph{considered} not repenting. The
counterfactuals are the text's own residual possibilities, not external
impositions.

\subsection{Ryan's Possible-Worlds Theory as Theoretical Backdrop}

A third framing is narratological. Ryan's possible-worlds theory
\cite{ryan} treats every narrative as already a multiverse: the
\textsc{taw} surrounded by the satellite worlds of every character's
beliefs, wishes, obligations, and intentions. Stories are textual
multiverses by their nature. Rukha makes the multiverse explicit and
computational, but it does not introduce the multiverse; it inherits it
from the text.

\subsection{What Rukha Will Not Do}

Rukha will not generate sexually explicit, violent, or theologically
inflammatory variants for shock value. The \textsc{transplant}
strategy in particular is gated: cross-tradition transplants will be
flagged and require explicit interpretive justification before
inclusion in the public corpus. The release will include explicit
interpretive framing language, consultation with religious-studies
scholars, and opt-in mechanisms for tradition-specific texts.

We acknowledge that this is not sufficient to address every possible
objection. We claim only that it is sufficient to do the work
honestly.

\diamondrule

\section{Related Work}

\subsection{Counterfactual Story Generation}

Qin et al.'s \textsc{TimeTravel} \cite{timetravel} is the foundational
single-step counterfactual rewriting dataset: 29,849 supervised rewrites
of \textsc{rocstories} narratives, with the empirical observation that
human rewriters change roughly 1.5 of 5 sentences. Chen et al.'s
\textsc{educat} \cite{educat} achieves unsupervised single-step
rewriting via causal-effect-localized minimal-edit estimation. The
\textsc{pasta} dataset \cite{pasta} introduces participant-state
counterfactual perturbations as a cleaner formalism than free-form
rewrites. Sketch-and-Customize \cite{sketchcustomize} provides a
two-stage architecture for counterfactual story generation. Chatzi et al.
\cite{chatzi} formalize counterfactual reasoning at the
token-generation level via Gumbel-Max \textsc{scm}s.

None of this work operates recursively. Each system performs a single
counterfactual against a canonical original. Rukha is the first system
to chain counterfactuals to depth $\geq 3$.

\subsection{Causal Reasoning in Language Models}

Kıcıman et al. \cite{kiciman} establish that language-model-built causal
graphs achieve high accuracy on benchmark causal-discovery tasks.
Vashishtha et al.'s \emph{Executable Counterfactuals}
\cite{executablecounterfactuals} decomposes counterfactual reasoning
into Pearl's three steps and reports that prior benchmarks routinely
\emph{skip the abduction step}, overestimating model performance.
Tang et al. \cite{tangcounterfactual} provide a decompositional study
across eleven datasets, with the cautionary finding that Tree-of-Thoughts
\emph{over-qualifies} on counterfactual reasoning by diverting the model
from contextually-supported causal pathways. \textsc{counterbench}
\cite{counterbench} is the canonical 2025 benchmark; its Nested subset
is the closest published analog to Rukha's depth-$\geq 2$ regime, and
the near-random baseline performance on Nested is the gap Rukha targets.

\subsection{Long-Form Narrative Consistency}

The long-form narrative-consistency literature has matured rapidly.
ConStory-Bench \cite{constory} establishes Consistency Error Density as
the canonical metric; NoCha \cite{nocha} establishes minimal-pair
True/False discrimination on book-length narratives;
NarrativeFactScore \cite{narrativefactscore} introduces the Character
Knowledge Graph as factuality oracle; SCORE \cite{scorestory} reports
substantial coherence improvements from explicit dynamic-state tracking;
\textsc{lostinstories} \cite{lostinstories} catalogs the empirical
failure modes (errors cluster in factual and temporal dimensions, around
the middle of long generations, in higher-entropy segments).

This literature defines the methodological terrain Rukha must operate
on. Our protocol (Section 6) inherits its metrics and adapts them to
recursion depth.

\subsection{Narrative Possible-Worlds Theory}

Ryan's \emph{Possible Worlds, Artificial Intelligence, and Narrative
Theory} \cite{ryan} is the canonical narratological framework. Lewis's
\emph{Counterfactuals} \cite{lewiscounterfactuals} grounds the modal
semantics. Doležel's \emph{Heterocosmica} and Pavel's \emph{Fictional
Worlds} provide complementary Prague-school readings. The 2019 Bell-Ryan
handbook \emph{Possible Worlds Theory and Contemporary Narratology}
gathers the recent state of the field.

The narratological tradition has long modeled narrative as a multiverse;
Rukha makes the multiverse computational.

\subsection{Industry and Patent Landscape}

A patent and grey-literature sweep covering USPTO, Google Patents, EPO,
and the major industrial research labs (Anthropic, OpenAI, AllenAI,
Microsoft Research, Google DeepMind, Meta FAIR, Cohere, Mistral, AI21,
Together) returned no system that combines Pearl-\textsc{aap}-grounded
narrative perturbation, recursive depth $\geq 3$, structural-sharing
infrastructure, and foundational-text seeding. The closest patents
(US 7{,}246{,}315 on personal narrative agents; US 10{,}657{,}201 on
configurable narrative generation; US 10{,}599{,}700 on narrative
detection) are genre-distinct.

The closest industry precedents are: AllenAI's
counterfactual-story-rewriting program (Qin and collaborators);
Microsoft Research's Pearl-\textsc{llm} causal program (Kıcıman,
Vashishtha); Google DeepMind's Genie 3 \cite{genie3} (counterfactual
\emph{visual} world models for agent training, distinct in modality);
and the MIT Media Lab thesis ``Next Week Tonight: Simulating
Counterfactual Narratives'' \cite{nextweektonight} (counterfactual
narrative simulation via \textsc{rag}-and-agents, architecturally
distinct from Pearl-do-calculus). Each addresses a component of Rukha;
none addresses the integrated whole.

\subsection{Cultural Attractors and Iterated Generation}

Perez et al. \cite{perezattractors} document that iterated language-model
transmission converges to attractor states for measurable
properties (toxicity, positivity, difficulty, length). The result is the
single most relevant empirical precedent for Rukha's
narrative-attractor hypothesis. The crucial distinction:
Perez et al. measure \emph{property-level} attractors over iterated
paraphrase chains; Rukha tests \emph{archetype-level} attractors over
structured perturbation graphs. The methodological inheritance is
explicit; the empirical question is distinct.

\diamondrule

\section{Limitations and Open Questions}

The paper makes a strong claim --- recursive counterfactual reasoning
at depth five and six on narrative event graphs is tractable with
appropriate scaffolding --- and the strength of the claim demands
forthright limitations.

\subsection{The Depth Claim Is Not Yet Empirically Demonstrated}

Section 6 specifies the evaluation protocol; we have not yet executed
it. The argument that depth five and six \emph{should} be tractable
under \textsc{aap} scaffolding rests on (a) Pearl-Robins decomposition
in acyclic \textsc{scm}s, (b) the structural-sharing storage argument,
and (c) the Reasoning Horizon and \textsc{counterbench} results
indirectly. None of these is the same as a demonstration. The paper
should be read as a research program with methodological commitments,
not as a results paper.

\subsection{Coherence Collapse Modes Are Not Fully Characterized}

We expect two failure modes at deep recursion: \emph{attractor
reversion}, where each recursion step pulls the narrative back toward
the canonical Jonah story (the language model's strongest prior), and
\emph{coherence drift}, where each rewrite introduces small
inconsistencies that compound. Our anti-attractor mechanism is
\textsc{xgrammar}-constrained \textsc{aap} prompting plus retrieval
anchoring against the perturbed-graph state. Whether this suffices for
depth six is the open empirical question.

\subsection{Load-Bearing Score Validation}

The composite load-bearing score $0.40 \cdot \text{BC} + 0.30 \cdot
\text{PR} + 0.30 \cdot \text{OD}$ is heuristic. The principled
alternative --- Halpern–Pearl actual causality \cite{halpernpearlcausality}
--- requires, for each candidate node $v$, computing the post-\textsc{aap}
graph-edit distance under perturbation. The composite score is a
topological proxy. Whether the proxy correlates with the principled
metric is testable; we will test it on the Jonah graph and report.

\subsection{Sacred-Text Counterfactuals Will Be Disorienting}

We have argued the case for computational midrash; we recognize that
the argument will not satisfy every reader. Some perturbations of Jonah
will produce trajectories that are theologically unsettling. We do not
think this is a reason not to do the work, but we also do not think it
is a reason to be careless about it. Public release will include
interpretive framing language, scholarly consultation, and opt-in
mechanisms for tradition-specific texts.

\subsection{Throughput and Compute Budget}

A full Rukha campaign on Jonah (depth six, branching factor five)
requires on the order of $5^6 = 15{,}625$ \textsc{aap}
forward-propagations, each producing a 16-event trajectory. With
4× NVIDIA \textsc{rtx pro} 6000 Blackwell cards (96\,GB \textsc{vram}
each, 384\,GB total) running Llama 3.3 70B via \textsc{vllm} tensor
parallelism, throughput is approximately 20--30 tokens per second per
replica. After deduplication via \textsc{faiss} cosine
similarity at the 0.92 threshold, the actual unique trajectory count is
expected to be substantially smaller than $15{,}625$ --- the
structural-sharing argument predicts compression ratios on the order of
$10^2$ to $10^3$ in this regime. Wall-clock for a full Jonah depth-six
campaign: weeks, not months. For longer texts (Mahābhārata, Iliad), the
sliding-context-with-causal-summary mechanism described in the original
manifesto applies.

\subsection{What Would Falsify Rukha}

The paper should specify what evidence would falsify its central claim.
Three findings would do so:

\begin{enumerate}
  \item Per-depth Consistency Error Density grows super-linearly with
    recursion depth, with a clear inflection between depth 3 and 4 ---
    indicating that the \textsc{aap} protocol is not in fact preserving
    abduction across recursion levels.
  \item Depth-six variants of Jonah do \emph{not} cluster in
    archetype space; instead they distribute uniformly. This would
    falsify the narrative-physics hypothesis.
  \item Depth-six variants are indistinguishable from depth-three
    variants under the NoCha minimal-pair protocol --- indicating that
    the model is treating recursion as a no-op, generating the same
    story with relabeled nodes.
\end{enumerate}

We commit to reporting these results regardless of outcome.

\diamondrule

\section{Coda}

In the multiverse of every story humanity ever told itself, there exist
trajectories where things went better. Where the sacrifice was not
required. Where the exile ended sooner. Where the prophet did not need
to be swallowed before he could be heard.

Rukha exists to make those trajectories visible. Not because we believe
the canonical text was wrong --- the canonical text is the trajectory
that survived, and there are good reasons it survived --- but because
the canonical text was always one trajectory among many, and what it
\emph{means} is partly defined by what it \emph{could have been}.

A counterfactual is not a hypothetical. It is a tool for reading. To
ask what would have happened if Jonah had obeyed is to read Jonah more
carefully than the canonical text alone allows. To ask the question
recursively, to depth five and six, is to read with the tradition's full
midrashic depth made computational and made shareable.

The output of a Rukha campaign is a corpus, yes. It is also a map of
narrative possibility, yes. But the deepest output is interpretive: it
is the topology of grace, the architecture of mercy made visible at
scale. Stories are how humans process the gap between what was and
what could have been. That gap is what mercy lives in. Rukha makes the
gap explorable.

\emph{Rukha doesn't just generate counterfactuals --- it generates
mercy.}

\vspace{1.4em}
\taperedrule
\vspace{0.4em}
\begin{center}\itshape\small
Returning books to breath.
\end{center}

\clearpage

% ─── Appendix A: Schema ──────────────────────────────────────────────
\appendix
\section*{\sceq{Appendix A.\ The Pydantic Schema}}
\addcontentsline{toc}{section}{Appendix A. The Pydantic Schema}

The event-extraction schema, expressed as Python dataclasses with
docstring guidelines per the GoLLIE / PyCode-TextEE convention:

\begin{small}\ttfamily\setstretch{1.05}
\noindent
@dataclass\\
class NarrativeEvent:\\
\hspace*{2em}id: str\hfill\# Unique ID, e.g. 'E01'\\
\hspace*{2em}trigger: str\hfill\# The word(s) triggering the event\\
\hspace*{2em}summary: str\hfill\# One-sentence description\\
\hspace*{2em}event\_type: str\hfill\# ACTION | STATE\_CHANGE | DECISION\\
\hspace*{2em}\hspace*{2em}\hspace*{6em}\hfill\# DIVINE\_ACTION | COMMUNICATION\\
\hspace*{2em}stac\_label: str\hfill\# Situation | Task | Action | Consequence\\
\hspace*{2em}agents: list[str]\hfill\# Characters involved\\
\hspace*{2em}location: str\hfill\# Where it occurs\\
\hspace*{2em}verse\_ref: str\hfill\# Source reference\\[0.4em]
@dataclass\\
class CausalEdge:\\
\hspace*{2em}source\_id: str\\
\hspace*{2em}target\_id: str\\
\hspace*{2em}causal\_type: str\hfill\# DIRECT\_CAUSE | ENABLES\\
\hspace*{2em}\hspace*{2em}\hspace*{6em}\hfill\# MOTIVATES | PREVENTS\\
\hspace*{2em}strength: float\hfill\# 0.0–1.0\\
\hspace*{2em}evidence: str\hfill\# Text span supporting this
\end{small}

\vspace{1em}

\section*{\sceq{Appendix B.\ Verifying the 39-Word Parallel}}
\addcontentsline{toc}{section}{Appendix B. Verifying the 39-Word Parallel}

The Magonet observation (39 Hebrew words in Jonah 4:2--3 and 4:10--11;
30 if maqqef-joined clusters are counted as single units) is
machine-verifiable against the Open Scriptures Hebrew Bible
\cite{oshb}, which provides the Westminster Leningrad Codex in OSIS
\textsc{xml} with each \texttt{<w>} element marking a graphic word. The
verification methodology, in pseudocode:

\begin{small}\ttfamily\setstretch{1.05}
\noindent
load oshb-jonah.xml\\
for verses in [(4,2), (4,3), (4,10), (4,11)]:\\
\hspace*{2em}words = xml.findall('.//w[@verse=verses]')\\
\hspace*{2em}graphic\_count = len(words)\\
\hspace*{2em}clusters = collapse\_maqqef\_joined(words)\\
\hspace*{2em}lexical\_count = len(clusters)\\
sum graphic\_count over (4,2) and (4,3): expect 39\\
sum graphic\_count over (4,10) and (4,11): expect 39\\
sum lexical\_count over (4,2) and (4,3): expect 30\\
sum lexical\_count over (4,10) and (4,11): expect 30
\end{small}

\vspace{0.6em}

\noindent
The published version of this paper will contain the script-derived
counts as primary evidence and will document any deviation from
Magonet's figures with the methodological reason
(maqqef-treatment edge cases, ketiv/qere variants, BHS apparatus
footnotes). The literary-critical observation remains Magonet's; the
machine verification establishes its mechanical reproducibility.

\vspace{1.4em}

% ─── Bibliography ────────────────────────────────────────────────────
\renewcommand{\refname}{\sceq{References}}
\addcontentsline{toc}{section}{References}

\begin{small}\setstretch{1.08}\setlength{\parskip}{0.4ex}

\begin{thebibliography}{99}

\bibitem{aaronsonwatrous}
S.\,Aaronson and J.\,Watrous.
\newblock Closed timelike curves make quantum and classical computing equivalent.
\newblock \emph{Proceedings of the Royal Society A} 465 (2009): 631–647.
\newblock arXiv:0808.2669.

\bibitem{bagwellhamt}
P.\,Bagwell.
\newblock Ideal hash trees.
\newblock EPFL technical report, 2001.

\bibitem{chatzi}
K.\,Chatzi et al.
\newblock Counterfactual token generation in large language models.
\newblock arXiv:2409.17027, 2024.

\bibitem{constory}
ConStory-Bench / ConStory-Checker.
\newblock Lost in stories: Consistency bugs in long story generation by \textsc{llm}s.
\newblock arXiv:2603.05890, 2026.

\bibitem{counterbench}
J.\,Chen et al.
\newblock CounterBench: Evaluating and improving counterfactual reasoning in large language models.
\newblock arXiv:2502.11008, 2025.

\bibitem{correaleebareinboim}
J.\,Correa, S.\,Lee, and E.\,Bareinboim.
\newblock Nested counterfactual identification from arbitrary surrogate experiments.
\newblock \emph{NeurIPS} 2021.
\newblock arXiv:2107.03190.

\bibitem{deepimagine}
DeepImagine: Learning biomedical reasoning via successive counterfactual imagining.
\newblock arXiv:2604.23054, 2026.

\bibitem{depthgeneralization}
Exploring depth generalization in large language models for solving recursive logic tasks.
\newblock arXiv:2512.02677, AAAI 2026.

\bibitem{driscolltarjan}
J.\,R.\,Driscoll, N.\,Sarnak, D.\,D.\,Sleator, and R.\,E.\,Tarjan.
\newblock Making data structures persistent.
\newblock \emph{Journal of Computer and System Sciences} 38 (1989): 86–124.

\bibitem{educat}
J.\,Chen et al.
\newblock \textsc{educat}: Unsupervised editing for counterfactual stories.
\newblock \emph{AAAI} 2022.
\newblock arXiv:2112.05417.

\bibitem{executablecounterfactuals}
A.\,Vashishtha et al.
\newblock Executable counterfactuals: Improving \textsc{llm}s' causal reasoning through code.
\newblock \emph{ICLR} 2026.
\newblock arXiv:2510.01539.

\bibitem{faithfulnessdecay}
Y.\,Ye et al.
\newblock Mechanistic evidence for faithfulness decay in chain-of-thought reasoning.
\newblock arXiv:2602.11201, 2026.

\bibitem{filliatreconchon}
J.-C.\,Filliâtre and S.\,Conchon.
\newblock Type-safe modular hash-consing.
\newblock \emph{ML Workshop}, 2006.

\bibitem{fivedbfmtt}
5D Brainfuck With Multiverse Time Travel.
\newblock \emph{Esolang Wiki}, esolangs.org, 2017.

\bibitem{freemanbetweenness}
L.\,C.\,Freeman.
\newblock A set of measures of centrality based on betweenness.
\newblock \emph{Sociometry} 40 (1977): 35–41.

\bibitem{genie3}
Google DeepMind.
\newblock Genie 3: A new frontier for world models.
\newblock Technical report, August 2025.

\bibitem{gollie}
O.\,Sainz, I.\,García-Ferrero, R.\,Agerri, O.\,Lopez de Lacalle, G.\,Rigau, and E.\,Agirre.
\newblock GoLLIE: Annotation guidelines improve zero-shot information-extraction.
\newblock \emph{ICLR} 2024.
\newblock arXiv:2310.03668.

\bibitem{halpernpearlcausality}
J.\,Y.\,Halpern.
\newblock \emph{Actual Causality}.
\newblock MIT Press, 2016.

\bibitem{ipfsmerkledag}
J.\,Benet.
\newblock IPFS — Content addressed, versioned, P2P file system.
\newblock arXiv:1407.3561, 2014.

\bibitem{kiciman}
E.\,Kıcıman, R.\,Ness, A.\,Sharma, and C.\,Tan.
\newblock Causal reasoning and large language models: Opening a new frontier for causality.
\newblock arXiv:2305.00050, 2023.

\bibitem{knuthzdd}
D.\,E.\,Knuth.
\newblock \emph{The Art of Computer Programming, Vol.~4, Fascicle 1: Bitwise Tricks \& Techniques; Binary Decision Diagrams}.
\newblock Addison-Wesley, 2009.

\bibitem{lewiscounterfactuals}
D.\,Lewis.
\newblock \emph{Counterfactuals}.
\newblock Blackwell, 1973.

\bibitem{lihybrid}
J.\,Li, Y.\,Pan, and X.\,Pi.
\newblock Beyond \textsc{llm}s: A linguistic approach to causal graph generation from narrative texts.
\newblock \emph{ACL} 2025 Workshop on Narrative Understanding.

\bibitem{lostinstories}
Lost in stories: consistency bugs in long story generation by \textsc{llm}s.
\newblock arXiv:2603.05890, 2026.

\bibitem{magonet}
J.\,Magonet.
\newblock \emph{Form and Meaning: Studies in Literary Techniques in the Book of Jonah}.
\newblock 2nd ed., Almond Press, Sheffield, 1983.

\bibitem{magonettorah}
J.\,Magonet.
\newblock Jonah's recalcitrant prayer.
\newblock TheTorah.com, 2014.

\bibitem{narrativefactscore}
NarrativeFactScore: Agent-as-a-judge for factual summarization of long narratives.
\newblock arXiv:2501.09993, 2025.

\bibitem{nextweektonight}
A.\,Gauri.
\newblock Next week tonight: Simulating counterfactual narratives.
\newblock SM-MAS thesis, MIT Media Lab, 2025.

\bibitem{nocha}
M.\,Karpinska et al.
\newblock One thousand and one pairs: A novel challenge for long-context language models.
\newblock \emph{EMNLP} 2024.
\newblock arXiv:2406.16264.

\bibitem{okasakipfds}
C.\,Okasaki.
\newblock \emph{Purely Functional Data Structures}.
\newblock Cambridge University Press, 1998.

\bibitem{oshb}
Open Scriptures Hebrew Bible (OSHB), version 2.0.
\newblock github.com/openscriptures/morphhb. Westminster Leningrad Codex in OSIS \textsc{xml}.

\bibitem{pagerank}
S.\,Brin and L.\,Page.
\newblock The anatomy of a large-scale hypertextual web search engine.
\newblock \emph{Computer Networks and ISDN Systems} 30 (1998): 107–117.

\bibitem{pasta}
M.\,Ghazarian, T.\,Wolfson et al.
\newblock \textsc{pasta}: A dataset for modeling participant states in narratives.
\newblock \emph{EMNLP} 2022.
\newblock arXiv:2208.00329.

\bibitem{pearlcausality}
J.\,Pearl.
\newblock \emph{Causality: Models, Reasoning, and Inference}.
\newblock 2nd ed., Cambridge University Press, 2009.

\bibitem{pearlalgorithmization}
J.\,Pearl.
\newblock The algorithmization of counterfactuals.
\newblock UCLA Technical Report R-360, 2013.

\bibitem{pearlrobins}
J.\,Pearl and J.\,Robins.
\newblock Probabilistic evaluation of sequential plans from causal models with hidden variables.
\newblock \emph{UAI} 1995.

\bibitem{perezattractors}
J.\,Perez et al.
\newblock When \textsc{llm}s play the telephone game: Cultural attractors as conceptual tools to evaluate \textsc{llm}s in multi-turn settings.
\newblock arXiv:2407.04503, 2024.

\bibitem{proppmorphology}
V.\,Propp.
\newblock \emph{Morphology of the Folktale}.
\newblock Russian original 1928; English translation, University of Texas Press, 1968.

\bibitem{pycodetexteed}
S.\,Srivastava et al.
\newblock PyCode-TextEE: Instruction-tuning \textsc{llm}s for event extraction with annotation guidelines.
\newblock \emph{ACL Findings} 2025.
\newblock arXiv:2502.16377.

\bibitem{reagansixarcs}
A.\,J.\,Reagan, L.\,Mitchell, D.\,Kiley, C.\,M.\,Danforth, and P.\,S.\,Dodds.
\newblock The emotional arcs of stories are dominated by six basic shapes.
\newblock \emph{EPJ Data Science} 5 (2016): 31.

\bibitem{ryan}
M.-L.\,Ryan.
\newblock \emph{Possible Worlds, Artificial Intelligence, and Narrative Theory}.
\newblock Indiana University Press, 1991.

\bibitem{sasson}
J.\,M.\,Sasson.
\newblock \emph{Jonah: A New Translation with Introduction, Commentary, and Interpretation}.
\newblock Anchor Bible 24B, Doubleday, 1990.

\bibitem{scorestory}
SCORE: Story coherence and retrieval enhancement for AI narratives.
\newblock arXiv:2503.23512, 2025.

\bibitem{sketchcustomize}
B.\,Hao, X.\,Pang et al.
\newblock Sketch and customize: A counterfactual story generator.
\newblock \emph{AAAI} 2021.

\bibitem{stepflowbreaks}
Reasoning fails where step flow breaks.
\newblock arXiv:2604.06695, 2026.

\bibitem{tangcounterfactual}
W.\,Tang et al.
\newblock On the eligibility of \textsc{llm}s for counterfactual reasoning: A decompositional study.
\newblock arXiv:2505.11839, 2025.

\bibitem{timetravel}
L.\,Qin, A.\,Bosselut, A.\,Holtzman, C.\,Bhagavatula, E.\,Clark, and Y.\,Choi.
\newblock Counterfactual story reasoning and generation.
\newblock \emph{EMNLP} 2019.
\newblock arXiv:1909.04076.

\bibitem{trible}
P.\,Trible.
\newblock \emph{Rhetorical Criticism: Context, Method, and the Book of Jonah}.
\newblock Fortress, 1994.

\bibitem{virtualhistory}
N.\,Ferguson, ed.
\newblock \emph{Virtual History: Alternatives and Counterfactuals}.
\newblock Macmillan, 1997.

\bibitem{xgrammar}
Y.\,Dong et al.
\newblock \textsc{xgrammar}: Flexible and efficient structured generation.
\newblock arXiv:2411.15100, 2024.

\end{thebibliography}
\end{small}

\end{document}
