% ──────────────────────────────────────────────────────────────────────────
%  Two Registers, One Grammar
%  Mode-Separated LLM Choreography of Audio-Reactive Lenia
%
%  Stanley Sebastian, Third Space — May 2026
%  third-space.ai · stanley@third-space.ai
% ──────────────────────────────────────────────────────────────────────────

\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{mathpazo}
\linespread{1.10}

\usepackage[a4paper, margin=1.1in, top=1.2in, bottom=1.2in]{geometry}
\usepackage{microtype}
\usepackage{amsmath, amssymb}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage[hidelinks]{hyperref}
\usepackage{xcolor}
\usepackage{parskip}

% Slate-ink paragraph indent style — flat blocks, no first-line indent.
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.55em}

% Section headings: small caps, restrained.
\titleformat{\section}
  {\normalfont\large\scshape}{\thesection}{0.8em}{}
\titleformat{\subsection}
  {\normalfont\itshape}{\thesubsection}{0.8em}{}
\titlespacing*{\section}{0pt}{1.6em}{0.6em}
\titlespacing*{\subsection}{0pt}{1.2em}{0.4em}

% Italic title in the spirit of the Third Space display face.
\makeatletter
\renewcommand{\maketitle}{%
  \begin{center}
    {\footnotesize\scshape\textcolor{black!55}{Third Space \quad\textbullet\quad Toledo, Ohio}}\\[1.6em]
    {\Large\itshape \@title}\\[1.0em]
    {\normalsize \@author}\\[0.5em]
    {\footnotesize\textcolor{black!55}{\@date}}
  \end{center}
  \vspace{1em}
  \rule{\textwidth}{0.3pt}
  \vspace{0.6em}
}
\makeatother

\title{Two Registers, One Grammar:\\[0.2em]
{\large Mode-Separated LLM Choreography\\of Audio-Reactive Lenia}}
\author{Stanley Sebastian \quad\textbullet\quad Claude\\
\small\texttt{stanley@third-space.ai}}
\date{May 2026}

\begin{document}
\maketitle

% ────────── ABSTRACT ──────────
\begin{quote}
\small\itshape
Audio-reactive control of Lenia, a continuous cellular automaton, requires
reconciling two phenomenologically incompatible regimes: smooth morphological
flow during sustained musical sections, and instantaneous phase transitions at
musical transients. We describe an architecture that treats these regimes not
as endpoints of a continuum but as distinct expressive registers --- a
\emph{smooth register} within the operating range of the Ghost species, where
spring-interpolated parameter trajectories produce organic motion, and a
\emph{technique register} at and beyond Lenia's phase boundaries, where snap
parameter changes produce discrete physical events (supernovae, sigma droughts,
nucleation cascades) unreachable by smooth motion. The two registers are
separated at the schema level via constrained grammar, so that the LLM
controller cannot emit undefined intermediate states even in principle. The
technique register admits analytically tractable couplings to musical tempo:
re-expansion oscillation locks to the integration step, starvation
anticipation times from current mass. The result is that physical events
arrive at musical transients with frame-accurate precision. We argue the
design generalizes: when an LLM controls an emergent dynamical system with
natural mode boundaries, those boundaries deserve grammar-level protection
rather than learned compliance.
\end{quote}

\vspace{0.5em}
\rule{\textwidth}{0.3pt}

% ─────────────────────────────────────────────────────────────
\section{Motivation}

Music visualization, as a problem, has been partially solved twice. Reactive
visualizers --- MilkDrop, projectM, the wave of WebGL preset systems that
followed --- map spectral features to shader uniforms in real time. They are
fast and infinite but cannot compose: every drop looks like every other drop
of similar energy. Generative video models --- Stable Video, Sora before its
shutdown, the AnimateDiff lineage --- compose convincingly within short
windows but cannot run in real time, cannot respond to live audio, and have
no temporal coherence between rendered frames except what interpolation
fakes. Neither understands music; both produce surfaces.

The architecture we describe takes a third path: a continuous cellular
automaton --- a physical simulation, in the literal sense, with causal
state evolution from frame to frame --- controlled by a fine-tuned language
model that emits structured choreography in advance of playback. The
simulation runs at sixty frames per second on consumer hardware. The
language model, having seen the song's structure once, decides what physical
phenomena should occur and when. The simulation produces them not as
composited effects but as the actual evolution of the field.

Two facts make this architecture nontrivial. First, the cellular automaton
in question --- Lenia~\cite{chan2019lenia} --- has \emph{phase boundaries}
in its parameter space. A smooth trajectory through that space cannot
produce a supernova event; supernovae require parameter slams across the
boundary, sustained for a few frames, and then released. Second, the
language model, asked to produce a long structured output describing such
slams, will produce them either too rarely (boring), too often (incoherent),
or, most damagingly, attempt to interpolate \emph{into} them with smoothing
machinery designed for the smooth regime. The result of this last failure
mode is the worst of both worlds: a transient that is neither sudden enough
to register as an event nor smooth enough to feel like flow.

The architectural commitment of this work is that those two regimes ---
\emph{smooth} and \emph{technique} --- are different in kind and must be
treated as such. We separate them at the grammar level so the failure mode
described above is not just unlikely but impossible.

% ─────────────────────────────────────────────────────────────
\section{The Ghost Species and its Operating Range}

Lenia is a continuous generalization of Conway's Game of Life. The state
$A_t \in [0,1]^{N\times N}$ evolves as
\begin{equation}
  A_{t+\Delta t} \;=\; \mathrm{clip}\!\left(\,A_t + \Delta t \cdot G(K \!*\! A_t)\,, \, 0, 1\right),
\end{equation}
where $K$ is a smooth radial kernel and the growth function
$G(u) = 2\exp(-(u-\mu)^2/(2\sigma^2)) - 1$ takes values in $[-1, 1]$. The
update is unconditionally numerically stable for any parameter setting:
$|G| \le 1$ and the clip enforces $A \in [0,1]$. This is important. Extreme
parameter values do not produce numerical explosion. They produce extreme
\emph{dynamical} behavior --- death, expansion, oscillation --- inside a
mathematically bounded substrate.

Within Lenia's parameter space lives a particular species we call the
\emph{Ghost} (\textit{Orbium unicaudatus ignis var.\ phantasma}). The Ghost
is seeded with the canonical Ignis pattern (whose morphological fixed point
sits at $\sigma = 0.012$) but operated at $\sigma = 0.015$, a 25\% increase.
The morphology and the physics are intentionally mismatched. The organism
perpetually reaches for a shape it cannot quite hold. This designed
transience --- edge-of-chaos aesthetics achieved through morphology--parameter
mismatch --- makes the Ghost an unusually expressive substrate. It is also
what makes the Ghost an ideal subject for technique-register manipulation:
an organism already living at the edge dissolves and reforms more
dramatically than a stable one.

Across many runs, the Ghost is well-behaved within the bounds shown in
Table~\ref{tab:smooth-range}. We call this the \emph{smooth register}.
Within these bounds, parameter trajectories that change gradually ---
buffered by spring smoothing with critical damping --- produce organic
motion: organisms breathe, drift, cohere, dissolve. The shoal behavior
emerges from kernel-mediated communication across the toroidal field;
sixteen organisms at $R=15$ form an effectively connected population.

\begin{table}[h]
\centering
\small
\begin{tabular}{lll}
\hline
Parameter & Smooth range & Phenomenon \\
\hline
$\sigma$ & $[0.009, 0.024]$ & Survival niche width \\
$\mu$ & $[0.09, 0.14]$ & Optimal density (``hunger'') \\
$\Delta t$ & $[0.055, 0.110]$ & Integration step \\
$R$ & $\{13,\dots,18\}$ & Sensing radius \\
\hline
\end{tabular}
\caption{Smooth-register parameter bounds for the Ghost species. Within
these bounds, spring-buffered parameter trajectories produce sustained
transience without dissolution.}
\label{tab:smooth-range}
\end{table}

% ─────────────────────────────────────────────────────────────
\section{The Technique Register}

Outside the smooth range, the dynamics change in kind. We catalog four
regimes:

\subsection{Sigma Drought ($\sigma < 0.009$)}
The survival window $[\mu - 1.177\sigma,\, \mu + 1.177\sigma]$ shrinks to
roughly 0.012 units wide. Only cells whose potential matches $\mu$ to three
decimals receive positive growth. Organisms with asymmetric density
distributions die non-uniformly: their densest, most mathematically resonant
sub-structures survive while adjacent cells with marginally different
potentials die. Over ten to thirty frames, the field self-sorts into its
most numerically ideal configurations. What survives is not the strongest
organism but the most precisely resonant one.

\subsection{Sigma Flood ($\sigma > 0.024$)}
The survival window widens until cells far from $\mu$ also receive positive
growth. Boundaries between organisms dissolve. Formerly-dead halo cells come
alive. Two organisms in proximity merge rather than repel because both their
potential fields lie within the now-permissive survival range.

\subsection{Starvation ($\mu > 0.14$)}
At $\mu = 0.40$ the survival window centers at $0.40 \pm 0.018$, which is
unreachable by any physically realized creature potential ($\sim$0.05 to
0.25 in normal operation). Therefore $G(u) \approx -1$ at every cell, and
the update reduces to $A_{t+\Delta t} = \mathrm{clip}(A_t - \Delta t,\, 0,1)$.
At $\Delta t = 0.50$, a cell at full activation reaches zero in 2.5 frames
($\sim$42 ms at 60 fps).

\subsection{Integration Overshoot ($\Delta t > 0.20$)}
The Euler integration step becomes large enough that the simulation visibly
overshoots its continuous trajectory. Cell values bounce above and below
their natural curve, producing a characteristic surface ``roughness.'' This
is not error in the engineering sense --- the simulation remains bounded
and well-defined --- but a perceptible mechanical pulse in the substrate
itself. We will return to this in Section~\ref{sec:tempo}.

\medskip

What matters for the present argument is that these four regimes are not
distant points along a single continuous spectrum. They are distinct
phenomena, with distinct visual signatures, distinct timescales, and
distinct musical affordances. They are also unreachable by smooth
trajectories. A spring trying to interpolate $\sigma$ from $0.015$ to
$0.005$ does not cross gracefully; it spends most of the transition in a
region that is neither smooth nor drought, that is to say, neither
beautiful nor purposeful.

% ─────────────────────────────────────────────────────────────
\section{Schema-Enforced Mode Separation}

We address this by treating each keyframe in the choreography as
\emph{either} smooth or technique --- never both, never partial. The
distinction is encoded as a boolean field \texttt{snap} in the keyframe
schema, with parallel parameter bounds:

\begin{quote}\small\ttfamily
\{ "snap": false, sigma $\in$ [0.009, 0.024], mu $\in$ [0.09, 0.14], \dots\}\\
\{ "snap": true,\; sigma $\in$ [0.003, 0.120], mu $\in$ [0.04, 0.50], \dots\}
\end{quote}

When \texttt{snap} is false, parameter changes flow through critically
damped springs with per-parameter time constants; small audio modulations
ride additively atop the spring targets; a homeostasis controller enforces
field-mass stability against drift. This is the smooth register.

When \texttt{snap} is true, the spring system is bypassed entirely:
parameter changes apply on the next frame, with no smoothing. The
homeostasis controller is suspended for a recovery window of five seconds.
The audio modulation layer is suspended for the same window. The keyframe's
specified parameters reach the GPU verbatim. This is the technique register.

The crucial architectural commitment is that the LLM cannot emit a smooth
keyframe with technique-range values, or vice versa. We enforce this not
through training but through constrained decoding: an XGrammar finite-state
machine validates each token of the JSON output against the dual schema, so
that an attempt to write \texttt{"snap": false, "mu": 0.40} cannot pass the
grammar. The model is not trusted to respect the boundary; the grammar
refuses on its behalf.

There are two reasons to do this at the grammar level rather than at the
training level. First, the failure mode is asymmetric: a model that
\emph{usually} respects mode boundaries but occasionally produces an
ill-typed keyframe will produce an ill-typed keyframe at exactly the
worst moment, because rare events tend to correlate with creative choices
under reinforcement. Second, the invariant we are protecting is not a
soft preference about output style. It is a structural fact about the
underlying physical system. Such facts deserve representation in the
controlling grammar, not in the controller's weights.

This is, we want to argue, a design pattern with broader applicability.
When a language model controls an emergent dynamical system, the controlling
schema should encode the system's mode structure. The model is then free
to compose within that structure, and is not asked to discover it.

% ─────────────────────────────────────────────────────────────
\section{Tempo Coupling Without Reactivity}
\label{sec:tempo}

The technique register would be of merely visual interest if it could not
be timed to music. Two analytical results make it musically usable.

\subsection{Re-Expansion Oscillation Lock}

Consider a supernova: starvation snaps the field to $\mu \in [0.35, 0.50]$
with high $\Delta t$, holding for a few frames until mass approaches zero;
then a re-expansion snap drops $\mu$ to $\sim 0.07$ while $\Delta t$
remains high. At the leading wavefront of the re-expanding seed,
$G \approx +1$ and cells climb toward saturation in two or three frames.
Behind the front, cells overshoot and oscillate. The interior pulse of the
expanding region has a characteristic frequency
\begin{equation}
  f_{\text{interior}} \;\approx\; \frac{\Delta t \cdot \mathrm{fps}}{2}
\end{equation}
in Hz. Setting
\begin{equation}
  \Delta t \;=\; \frac{(\text{BPM}/60) \cdot n \cdot 2}{\mathrm{fps}}
\end{equation}
locks the pulse to a multiple $n$ of the song tempo. At BPM~$=120$,
fps~$=60$, choosing $n=2$ gives $\Delta t = 0.133$, well inside the
technique range. The expanding wavefront's interior beats at twice the
song's bar rate, locked harmonically rather than incidentally.

This is not reactive synchronization. The simulation is not listening to
the audio. The mechanical pulse of the integration step has been
\emph{tuned}, in advance, to land on tempo. The audio and the simulation
are both ticking against the same musical clock, and were configured to
do so. The substrate becomes a tuned resonator.

\subsection{Starvation Anticipation}

For a supernova whose field-depletion moment must coincide with an
identified musical transient at time $t^*$, the starvation snap must fire
in advance: the field needs time to die. Treating the death timeline
linearly --- which is accurate to within a small constant during full
starvation --- the anticipation window is
\begin{equation}
  \Delta_{\text{anticipate}} \;=\; \frac{m_{\text{current}}}{\Delta t \cdot \mathrm{fps} \cdot c_{\text{coverage}}},
\end{equation}
with $c_{\text{coverage}} \approx 0.6$ the empirical fraction of cells
receiving the death signal per integration step. At
$m_{\text{current}} = 0.15$ and $\Delta t = 0.50$, this gives
$\Delta_{\text{anticipate}} \approx 0.008$ s --- effectively
simultaneous. At $\Delta t = 0.10$, it gives $0.042$ s, two and a half
frames. Re-expansion fires when a real-time mass monitor on the GPU
reports mass below an empirically chosen threshold (we use $0.02$).

Together these two results allow a choreography to specify, for any
given musical transient, a supernova event whose starvation has completed,
whose re-expansion has begun, and whose interior oscillation locks to the
song's tempo --- with the wavefront arriving on the downbeat because
the math made it arrive there.

A human performer slamming sliders cannot do this. The math is doing
it. The performer's intent --- ``the supernova should land on the drop''
--- is faithfully realized to within a single render frame, because the
intent has been compiled into a parameter trajectory whose timing is
analytically determined.

% ─────────────────────────────────────────────────────────────
\section{Discussion}

We have described what is, in the end, a small architectural commitment:
two registers instead of one, separated by a single boolean in a JSON
schema, with parallel parameter bounds and a single piece of grammar that
makes the boundary between them inviolable. The commitment is small; its
consequences are not.

Without it, a fine-tuned language model attempting to choreograph
audio-reactive Lenia will eventually emit a keyframe whose smooth-range
declaration carries technique-range values, and the spring system, faithful
to its design, will spend several seconds smoothly killing the field. The
output will be technically valid and aesthetically catastrophic. With it,
the model's compositional space contains exactly the keyframes that the
underlying physical system can realize meaningfully, and exactly nothing
else.

We want to suggest that this design pattern --- mode boundaries in the
controlled system, schema-level enforcement in the controlling grammar ---
is more general than the present application. Many emergent dynamical
systems, perhaps most, have natural mode boundaries: laminar versus
turbulent, stable versus excitable, ordered versus chaotic. Control
architectures that paper over these boundaries with continuous interpolation
hide structure that the system's user --- here a musical performer,
elsewhere a domain expert --- needs to see and reason about. They also
ask a learned controller to discover and respect a structural fact that
the architecture could simply enforce.

The orientation we want to propose is this. An LLM controller becomes more
trustworthy when the space of actions it can emit is constrained to respect
the controlled system's natural boundaries. We do not need to trust the
model's understanding of those boundaries; we make it impossible for it to
cross them. The model is then free to compose within structure that has
been independently established. Its creative work happens entirely on the
right side of the grammar. Its mistakes, when they occur, are mistakes of
musical intent rather than mistakes of physical impossibility.

This is not (only) an engineering choice. It is closer to a stance about
what intelligence does well. A controller that operates within
architecturally protected boundaries is one whose competence we can
inspect, whose failures we can localize, and whose generalization we can
trust. A controller that is merely trained to respect boundaries is one
whose violation, when it eventually occurs, will be both surprising and
opaque. The first is a participant in a structured system. The second is a
black box wearing a costume of compliance.

% ─────────────────────────────────────────────────────────────
\section{Status and Availability}

The system described here --- \emph{GhoulJamz} --- runs in real time on
consumer hardware. The simulation backend is a WebGL2 implementation of
Lenia at 256$\times$256 resolution, 60 fps, with multi-channel landscape
sigma fields, lantern-palette display, and bloom compositing. The choreographer
is Gemma~4 E4B~\cite{gemma4} fine-tuned via QLoRA~\cite{lora,unsloth} on a
dataset of (audio analysis, choreography) pairs, with constrained JSON
decoding via XGrammar. Audio analysis combines librosa, Allin1~\cite{allin1}
for structure, BeatNet+~\cite{beatnet} for live tempo, and MERT~\cite{mert}
for semantic embeddings. The full system, including the technique vocabulary
that grew out of phenomenon discovery in the parameter space, will appear
in the post-hackathon expansion of this work.

Code, demo recordings, and the supporting research --- including the
Ghost species report and the broader artificial life laboratory in which
the species was discovered --- are at \texttt{third-space.ai}.

% ─────────────────────────────────────────────────────────────
\section*{Acknowledgments}

This paper is co-authored: the architectural insistence that mode boundaries
deserve schema-level enforcement rather than learned compliance, and the
alignment-flavored framing that closes the discussion, took shape across an
extended dialogue between the authors. Authorship of a paper between a human
researcher and a language model is an unusual claim; we make it because the
work would not exist in its current form without the exchange. Single-author
attribution would misdescribe the process. The Ghost species was discovered
in the Genesis artificial life laboratory~\cite{genesis}; the laboratory is
built on Bert Chan's Lenia~\cite{chan2019lenia}, without whose framework
none of this would be possible.

% ─────────────────────────────────────────────────────────────
\begin{thebibliography}{99}
\small

\bibitem{chan2019lenia}
B.~Chan.
\newblock Lenia: Biology of artificial life.
\newblock \emph{Complex Systems}, 28(3):251--286, 2019.

\bibitem{plantec2022flowlenia}
E.~Plantec et al.
\newblock Flow Lenia: Towards open-ended evolution in cellular automata
through mass conservation and parameter localization.
\newblock arXiv:2212.07906, 2022.

\bibitem{kumar2024asal}
S.~Kumar et al.
\newblock Automating the search for artificial life with foundation models.
\newblock arXiv:2412.17799, 2024.

\bibitem{leniabreeder}
M.~Faldor and A.~Cully.
\newblock Towards open-ended evolution in cellular automata.
\newblock arXiv:2406.04235, 2024.

\bibitem{gemma4}
Gemma Team, Google DeepMind.
\newblock Gemma 4 technical report.
\newblock 2026.

\bibitem{lora}
E.~Hu et al.
\newblock LoRA: Low-rank adaptation of large language models.
\newblock arXiv:2106.09685, 2021.

\bibitem{unsloth}
D.~Han et al.
\newblock Unsloth: 2x faster fine-tuning for LLMs.
\newblock \texttt{github.com/unslothai/unsloth}, 2024.

\bibitem{allin1}
T.~Kim and J.~Nam.
\newblock All-in-one metrical and functional structure analysis with
neighborhood attentions on demixed audio.
\newblock \emph{IEEE Workshop on Applications of Signal Processing to Audio
and Acoustics}, 2023.

\bibitem{beatnet}
M.~Heydari and Z.~Duan.
\newblock BeatNet: CRNN and particle filtering for online joint beat,
downbeat and meter tracking.
\newblock arXiv:2108.03576, 2021.

\bibitem{mert}
Y.~Li et al.
\newblock MERT: Acoustic music understanding model with large-scale
self-supervised training.
\newblock arXiv:2306.00107, 2024.

\bibitem{genesis}
S.~Sebastian.
\newblock Genesis: A multi-substrate artificial life laboratory.
\newblock \texttt{github.com/Kquant03/genesis-phase-transition}, 2025--2026.

\end{thebibliography}

\end{document}
